---
title: "Unveiling wildfire patterns in Catalonia (1998-2022)"
author: "Gemma Bargalló Solé"
date: "`r Sys.Date()`"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme: cosmo
    vertical_layout: fill
---

```{r warning=FALSE, message=FALSE}
# 1. packages import
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(nanoparquet)
library(here)
library(scales)
library(naniar) 
library(skimr) 
library(patchwork)
library(VIM)
library(gt)
library(purrr)
library(forcats)
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
library(leaflet)
library(dplyr)
library(crosstalk)
library(shiny)
library(shinydashboard)  

# 2. Data Import
wfc_final2 <- read_csv(here("Data", "wfc_final2.csv"))

```

# 1. Introduction: The fire challenge in Catalonia

Wildfires are one of the most significant environmental threats in Catalonia. As climate change intensifies and traditional land management fades, the Mediterranean landscape becomes increasingly vulnerable.

![Source: [El Periódico (2021). Wildfire risk factors in Catalonia.](https://www.elperiodico.cat/ca/successos/20210709/tres-claus-risc-incendis-forestals-11897390)](images/incendis%20catalunya.jpg){width="100%"}

## The objectives of this visualization:

-   **Identify spatio-temporal patterns**: To understand not just when it burns, but where the recurrence is highest.

-   **Correlate causes and landscape**: To see how human activity and land cover types interact to create risk.

-   **Bridge descriptive and predictive analysis**: To provide a visual foundation that justifies the variables used in our Random Forest predictive model.

"*Data is the first step toward effective prevention. By visualizing the past 25 years, we can better prepare for the next season.*"

```{r}
# In this first section, we can show a summary "Value Box" 
# to show the total magnitude of the dataset.
total_fires <- nrow(wfc_final2)
total_area <- sum(wfc_final2$superficie_total_forestal, na.rm = TRUE)

# Custom display using valueBoxes
fluidRow(
  valueBox(total_fires, subtitle = "Total Fires Recorded", icon = icon("fire"), color = "red"),
  valueBox(round(total_area / 1000, 1), subtitle = "Total Thousand Ha Burnt", icon = icon("tree"), color = "yellow")
)
```

# 2. Chapter 1: The pulse of time (temporal analysis)

### Understanding 25 years of Fire History {data-commentary-width="350"}

```{r warning=FALSE, message=FALSE}
# 2. Data Preparation
# 2.1. Yearly Data: Convert year to numeric to avoid factor errors in scales
yearly_counts <- wfc_final2 %>%
  count(year) %>%
  mutate(year = as.numeric(as.character(year)))

# 2.2. Monthly Data: Define chronological order for the factor
wfc_final2$month_detected <- factor(wfc_final2$month_detected, 
                                    levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

monthly_counts <- wfc_final2 %>%
  count(month_detected)

# 3. Create Plots

# Plot A: Yearly Evolution with Trend Line
p_year <- ggplot(yearly_counts, aes(x = year, y = n)) +
  # Add the trend line (Linear Regression)
  geom_smooth(method = "lm", color = "steelblue", linetype = "dashed", size = 0.8, se = FALSE) +
  # Keep the actual data points and lines
  geom_line(color = "darkred", size = 1) +
  geom_point(color = "darkred", size = 2) +
  # Force X-axis to show every single year vertically
  scale_x_continuous(breaks = seq(min(yearly_counts$year), max(yearly_counts$year), by = 1)) +
  theme_minimal() +
  labs(title = "Annual fire frequency evolution", 
       subtitle = "1998 - 2022 Historical Series with Linear Trend",
       x = "Year", 
       y = "Number of Fires") +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1) # Vertical labels
  )

# Plot B: Monthly Seasonality (Bar Chart)
p_month <- ggplot(monthly_counts, aes(x = month_detected, y = n, group = 1)) +
  geom_bar(stat = "identity", fill = "indianred3") +
  theme_minimal() +
  labs(title = "Monthly fire seasonality", 
       subtitle = "Aggregated frequency (1998-2022)",
       x = "Month", 
       y = "Number of fires") +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# 4. Final Layout Assembly
# Side-by-side distribution using patchwork
final_temporal_plot <- p_year + p_month + 
  plot_annotation(
    title = 'Temporal analysis of wildfires in Catalonia',
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5, face = "bold"))
  )

# 5. Display the result
print(final_temporal_plot)

```

To understand the current risk, we must look at the historical series (1998-2022). As stated in the literature:

- **Interannual variability**: Wildfires in Catalonia are episodic. Years like 1998, 2005, 2009 and 2012 represent critical thresholds of drought and heat.

- **The seasonal double peak**: While summer is the main threat, the secondary peak in March highlights the danger of agricultural winter fires.

- **A slow descent**: We observe a slight decrease in ignitions over time, reflecting more efficient detection and prevention measures.


# 3. Chapter 2: The core of the story (interactive exploration)

```{r}
# MAP
# 1. Data Preparation
wfc_map_data <- wfc_final2 %>%
  mutate(
    # Year MUST be numeric for the slider to work correctly
    year_num = as.numeric(as.character(year)),
    # Create the customized popup text in English
    popup_text = paste0(
      "<div style='font-family: Arial; font-size: 12px;'>",
      "<strong>Municipality:</strong> ", municipio, "<br>",
      "<strong>Year:</strong> ", year, "<br>",
      "<strong>Burnt Surface:</strong> ", superficie_total_forestal, " ha<br>",
      "<strong>Land Cover ID:</strong> ", land_cover_id, "<br>",
      "<strong>Cause:</strong> ", causa, "</div>"
    )
  ) %>%
  select(longitude, latitude, year_num, popup_text, superficie_total_forestal, municipio)

# 2. Create SharedData object
sd <- SharedData$new(wfc_map_data)

# 3. Interactive Layout with Slider
bscols(
  widths = c(12),
  # Year Slider
  filter_slider(
    id = "year_slider", 
    label = "Slide to navigate through years (1998-2022):", 
    sharedData = sd, 
    column = ~year_num,
    step = 1,          # Move year by year
    ticks = TRUE,      # Show year marks
    sep = "",          # Avoid commas in years (e.g., 2,022 -> 2022)
    width = "100%"
  ),
  # Leaflet Map
  leaflet(sd, width = "100%", height = 700) %>%
    addTiles() %>% 
    addProviderTiles(providers$CartoDB.Positron) %>%
    addCircleMarkers(
      lng = ~longitude, 
      lat = ~latitude,
      radius = ~sqrt(superficie_total_forestal) + 3, 
      color = "#E44D2E",
      stroke = TRUE,
      weight = 1,
      fillOpacity = 0.7,
      popup = ~popup_text,
      label = ~paste(municipio, "-", superficie_total_forestal, "ha")
    )
)
```

### Where does Catalonia burn?

This interactive map bridges the gap between raw data and territorial management:

- **Regional pressure**: You can visually confirm the high density of events in the Barcelona Metropolitan Area and the Empordà.

- **Magnitude vs. frequency**: Use the slider to see how certain years were dominated by many small fires, while others were defined by a few Large Forest Fires.

- **Spatial patterns**: This visualization justifies why Latitude and Longitude became the most important variables in our predictive model.


