---
title: "Unveiling wildfire patterns in Catalonia (1998-2022)"
author: "Gemma Bargalló Solé"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r warning=FALSE, message=FALSE}
# 1. packages import
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(nanoparquet)
library(here)
library(scales)
library(naniar) 
library(skimr) 
library(patchwork)
library(VIM)
library(gt)
library(purrr)
library(forcats)
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
library(leaflet)
library(dplyr)
library(crosstalk)
library(shiny)
library(shinydashboard)  
library(plotly)

# 2. Data Import
wfc_final2 <- read_csv(here("Data", "wfc_final2.csv"))

```

# 1. Introduction: The fire challenge in Catalonia

Wildfires are one of the most significant environmental threats in Catalonia. As climate change intensifies and traditional land management fades, the Mediterranean landscape becomes increasingly vulnerable.

![Source: [El Periódico (2021). Wildfire risk factors in Catalonia.](https://www.elperiodico.cat/ca/successos/20210709/tres-claus-risc-incendis-forestals-11897390)](images/incendis%20catalunya.jpg){width="100%"}

## The objectives of this visualization:

-   **Identify spatio-temporal patterns**: To understand not just when it burns, but where the recurrence is highest.

-   **Correlate causes and landscape**: To see how human activity and land cover types interact to create risk.

-   **Bridge descriptive and predictive analysis**: To provide a visual foundation that justifies the variables used in our Random Forest predictive model.

"*Data is the first step toward effective prevention. By visualizing the past 25 years, we can better prepare for the next season.*"

```{r}
# In this first section, we can show a summary "Value Box" 
# to show the total magnitude of the dataset.
total_fires <- nrow(wfc_final2)
total_area <- sum(wfc_final2$superficie_total_forestal, na.rm = TRUE)

# Custom display using valueBoxes
fluidRow(
  valueBox(total_fires, subtitle = "Total Fires Recorded", icon = icon("fire"), color = "red"),
  valueBox(round(total_area / 1000, 1), subtitle = "Total Thousand Ha Burnt", icon = icon("tree"), color = "yellow")
)
```

# 2. Chapter 1: The pulse of time (temporal analysis)

### Understanding 25 years of Fire History {data-commentary-width="350"}

```{r warning=FALSE, message=FALSE, fig.height=7, fig.width=10}
# 2. Data Preparation
# 2.1. Yearly Data: Convert year to numeric to avoid factor errors in scales
yearly_counts <- wfc_final2 %>%
  count(year) %>%
  mutate(year = as.numeric(as.character(year)))

# 2.2. Monthly Data: Define chronological order for the factor
wfc_final2$month_detected <- factor(wfc_final2$month_detected, 
                                    levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

monthly_counts <- wfc_final2 %>%
  count(month_detected)

# 3. Create Plots

# Plot A: Yearly Evolution with Trend Line
p_year <- ggplot(yearly_counts, aes(x = year, y = n)) +
  # Add the trend line (Linear Regression)
  geom_smooth(method = "lm", color = "steelblue", linetype = "dashed", size = 0.8, se = FALSE) +
  # Keep the actual data points and lines
  geom_line(color = "darkred", size = 1) +
  geom_point(color = "darkred", size = 2) +
  # Force X-axis to show every single year vertically
  scale_x_continuous(breaks = seq(min(yearly_counts$year), max(yearly_counts$year), by = 1)) +
  theme_minimal() +
  labs(title = "Annual fire frequency evolution", 
       subtitle = "1998 - 2022 Historical Series with Linear Trend",
       x = "Year", 
       y = "Number of Fires") +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1) # Vertical labels
  )

# Plot B: Monthly Seasonality (Bar Chart)
p_month <- ggplot(monthly_counts, aes(x = month_detected, y = n, group = 1)) +
  geom_bar(stat = "identity", fill = "indianred3") +
  theme_minimal() +
  labs(title = "Monthly fire seasonality", 
       subtitle = "Aggregated frequency (1998-2022)",
       x = "Month", 
       y = "Number of fires") +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# 4. Final Layout Assembly
# Side-by-side distribution using patchwork
final_temporal_plot <- p_year + p_month + 
  plot_annotation(
    title = 'Temporal analysis of wildfires in Catalonia',
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5, face = "bold"))
  )

# 5. Display the result
print(final_temporal_plot)
```

To understand the current risk, we must look at the historical series (1998-2022). As shown in the Annual Wildfire Summary table and the temporal plots:

-   **Interannual variability**: Wildfires in Catalonia are episodic. The table highlights that years like 1998 and 2012 are not just frequency peaks but true ecological disasters, with 20,992 ha and 15,026 ha burnt respectively. These represent critical thresholds of drought and heat.

-   **Frequency vs. severity:** By comparing the line chart with the table, we notice an interesting pattern: a high number of fires (like in 2005 with 894 ignitions) doesn't always correlate with the largest burnt area (5,495 ha), suggesting that fire suppression efficiency varies depending on meteorological synoptics.

-   **The seasonal double peak**: While the bar chart shows summer is the main threat, the secondary peak in March highlights the danger of agricultural winter fires, a factor to be further analyzed in the "Causes" section.

-   **A slow descent**: The blue trend line in the plot, supported by the lower figures in the last decade of the table (e.g., 2018 or 2020), reflects a slight decrease in ignitions over time. This trend suggests that prevention measures are mitigating the increasing climatic risk.

### Detailed annual data

```{r}
# 1. Prepare the data
table_data <- wfc_final2 %>%
  group_by(year) %>%
  summarise(
    Total_Fires = n(),
    Total_Area_Ha = sum(superficie_total_forestal, na.rm = TRUE)
  ) %>%
  arrange(year)

# 2. Create the corrected gt table
fire_table <- table_data %>%
  gt() %>%
  tab_header(
    title = md("**Annual wildfire summary**"),
    subtitle = "Frequency and Burnt Surface (1998-2022)"
  ) %>%
  cols_label(
    year = "Year",
    Total_Fires = "Number of Fires",
    Total_Area_Ha = "Total Area (ha)"
  ) %>%
  fmt_number(
    columns = c(Total_Fires, Total_Area_Ha),
    decimals = 1,
    use_seps = TRUE
  ) %>%
  # Updated styling method
  tab_style(
    style = cell_fill(color = "#F9F9F9"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.width = pct(100),
    column_labels.font.weight = "bold"
  )

fire_table %>%
  tab_options(
    container.height = px(400), 
    container.overflow.y = TRUE 
  )

```

# 3. Chapter 2: The core of the story (interactive exploration)

```{r}
wfc_final2 <- wfc_final2 %>%
  mutate(
    # Fem el canvi basant-nos en la fila (row_number())
    municipio = case_when(
      row_number() == 778 ~ "AGUILAR DE SEGARRA",
      row_number() == 779 ~ "CARDONA",
      row_number() == 2096 ~ "OLIVELLA",
      row_number() == 2386 ~ "LA SÈNIA",
      row_number() == 2919 ~ "VALLCLARA",
      row_number() == 3348 ~ "MONTMANEU",
      row_number() == 3516 ~ "TORDERA",
      row_number() == 3524 ~ "SENAN",
      row_number() == 4740 ~ "SERÒS",
      row_number() == 4793 ~ "MARGALEF",
      row_number() == 4797 ~ "CARDONA",
      row_number() == 6068 ~ "SANT SALVADOR DE TORROELLA",
      row_number() == 7287 ~ "CASTELLFOLLIT DE RIUBREGÓS",
      row_number() == 8810 ~ "EL PLA DE MANLLEU",
      TRUE ~ municipio 
    ),
    
    provincia = case_when(
      row_number() == 778 ~ "BARCELONA",
      row_number() == 779 ~ "BARCELONA",
      row_number() == 2096 ~ "BARCELONA",
      row_number() == 2386 ~ "TARRAGONA",
      row_number() == 2919 ~ "TARRAGONA",
      row_number() == 3348 ~ "BARCELONA",
      row_number() == 3516 ~ "BARCELONA",
      row_number() == 3524 ~ "TARRAGONA",
      row_number() == 4740 ~ "LLEIDA",
      row_number() == 4793 ~ "TARRAGONA",
      row_number() == 4797 ~ "BARCELONA",
      row_number() == 6068 ~ "BARCELONA",
      row_number() == 7287 ~ "BARCELONA",
      row_number() == 8810 ~ "TARRAGONA",
      TRUE ~ provincia
    ),
    
    county_clean = case_when(
      row_number() == 778 ~ "BAGES",
      row_number() == 779 ~ "BAGES",
      row_number() == 2096 ~ "GARRAF",
      row_number() == 2386 ~ "MONTSIÀ",
      row_number() == 2919 ~ "CONCA DE BARBERÀ",
      row_number() == 3348 ~ "ANOIA",
      row_number() == 3516 ~ "MARESME",
      row_number() == 3524 ~ "CONCA DE BARBERÀ",
      row_number() == 4740 ~ "SEGRIÀ",
      row_number() == 4793 ~ "PRIORAT",
      row_number() == 4797 ~ "BAGES",
      row_number() == 6068 ~ "BAGES",
      row_number() == 7287 ~ "ANOIA",
      row_number() == 8810 ~ "ALT CAMP",
      TRUE ~ county_clean
    )
  )
```

```{r}
# Creem el dataframe de referència (diccionari)
land_cover_lookup <- data.frame(
  land_cover_id = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 
                    19, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 30, 31, 32, 33, 
                    34, 35, 36, 37, 38, 39, 40, 41, 0),
  land_cover_type = c(
    "Herbaceous crops", "Market gardens, nurseries and greenhouse crops", 
    "Vineyards", "Olive groves", "Other woody crops", "Crops in transformation", 
    "Dense coniferous forests", "Dense deciduous broadleaf forests", 
    "Dense sclerophyllous and laurel forests", "Shrubland", "Open coniferous forests", 
    "Open deciduous broadleaf forests", "Open sclerophyllous and laurel forests", 
    "Grasslands", "Riparian forest", "Bare forest soil", "Burnt areas", 
    "Rocky areas and scree", "Beaches", "Wetlands", "Urban core", 
    "Urban expansion", "Low-density urban areas", "Isolated buildings in rural areas", 
    "Isolated residential areas", "Green urban areas", "Industrial, commercial and/or service areas", 
    "Sports and leisure areas", "Mining extraction areas and/or landfills", 
    "Areas under transformation", "Road network", "Bare urban soil", 
    "Airport areas", "Railway network", "Port areas", "Reservoirs", 
    "Lakes and lagoons", "Watercourses", "Ponds", "Artificial canals", "Sea", "No data"
  )
)
# Comprovem que land_cover_id sigui numeric en ambdós costats
wfc_final2$land_cover_id <- as.numeric(as.character(wfc_final2$land_cover_id))

# Afegim la descripció
wfc_final2 <- wfc_final2 %>%
  left_join(land_cover_lookup, by = "land_cover_id")
```

```{r}
# MAP
# 1. Data Preparation for Leaflet
wfc_map_data <- wfc_final2 %>%
  mutate(
    # Year MUST be numeric for the slider to work correctly
    year_num = as.numeric(as.character(year)),
    
    # Create the customized popup text in English
    popup_text = paste0(
      "<div style='font-family: Arial; font-size: 12px;'>",
      "<b style='color: #8B0000;'>Municipality:</b> ", municipio, "<br>",
      "<strong>Year:</strong> ", year, "<br>",
      "<strong>Detected:</strong> ", date_detected, "<br>",
      "<strong>Extinguished:</strong> ", date_extinguished, "<br>",
      "<strong>Burnt Surface:</strong> ", superficie_total_forestal, " ha<br>",
      "<strong>Land Cover Type:</strong> ", land_cover_type, "<br>",
      "<strong>Cause:</strong> ", causa, 
      "</div>"
    )
  ) %>%
  # Seleccionem les variables necessàries i traiem el land_cover_id
  select(longitude, latitude, year_num, popup_text, superficie_total_forestal, 
         municipio, date_detected, date_extinguished, land_cover_type)

# 2. Create SharedData object
sd <- SharedData$new(wfc_map_data)

# 3. Interactive Layout with Slider
bscols(
  widths = c(12),
  # Year Slider
  filter_slider(
    id = "year_slider", 
    label = "Slide to navigate through years (1998-2022):", 
    sharedData = sd, 
    column = ~year_num,
    step = 1,          # Move year by year
    ticks = TRUE,      # Show year marks
    sep = "",          # Avoid commas in years (e.g., 2,022 -> 2022)
    width = "100%"
  ),
  # Leaflet Map
  leaflet(sd, width = "100%", height = 700) %>%
    addTiles() %>% 
    addProviderTiles(providers$CartoDB.Positron) %>%
    addCircleMarkers(
      lng = ~longitude, 
      lat = ~latitude,
      radius = ~sqrt(superficie_total_forestal) + 3, 
      color = "#E44D2E",
      stroke = TRUE,
      weight = 1,
      fillOpacity = 0.7,
      popup = ~popup_text,
      label = ~paste(municipio, "-", superficie_total_forestal, "ha")
    )
)
```

### Where does Catalonia burn?

This interactive map bridges the gap between raw data and territorial management:

-   **Regional pressure**: You can visually confirm the high density of events in the Barcelona Metropolitan Area and the Empordà.

-   **Magnitude vs. frequency**: Use the slider to see how certain years were dominated by many small fires, while others were defined by a few Large Forest Fires.

-   **Spatial patterns**: This visualization justifies why Latitude and Longitude became the most important variables in our predictive model.

In addition to the interactive map, we have generated these bar charts to summarize the provinces and municipalities with the highest fire frequency and the most significant burnt area. This dual perspective allows us to distinguish between areas of high human or natural activity (frequency) and areas of high ecological vulnerability (burnt area).

```{r}
# 1. Preparació de dades (Top 10)
top_10_muni_freq <- wfc_final2 %>%
  count(municipio) %>%
  slice_max(n, n = 10)

top_10_muni_area <- wfc_final2 %>%
  group_by(municipio) %>%
  summarise(total_area = sum(superficie_total_forestal, na.rm = TRUE)) %>%
  slice_max(total_area, n = 10)

area_provincia <- wfc_final2 %>%
  group_by(provincia) %>%
  summarise(total_area = sum(superficie_total_forestal, na.rm = TRUE))

# Paleta de blaus
blue_palette <- c("#08306b", "#08519c", "#2171b5", "#4292c6")

# Funció d'estil optimitzada per a gràfics horitzontals
estil_tfm_horizontal <- function() {
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 11, color = "#2c3e50"),
    axis.text.y = element_text(size = 9), # Noms de municipis més clars
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank(), # Treiem graella horitzontal
    panel.grid.minor = element_blank(),
    plot.margin = margin(5, 15, 5, 5)
  )
}

# --- FREQUENCY PLOTS (HORIZONTALS) ---

# P1: Província Freq
p1 <- ggplot(wfc_final2, aes(y = reorder(provincia, provincia, function(x) length(x)), x = ..count..)) +
  geom_bar(fill = blue_palette[1], width = 0.7) +
  geom_text(stat='count', aes(label=..count..), hjust=-0.2, size=3, fontface="bold") +
  expand_limits(x = count(wfc_final2, provincia) %>% pull(n) %>% max() * 1.2) +
  labs(title = "Fire frequency by province", y = NULL, x = "Number of fires") +
  estil_tfm_horizontal()

# P2: Municipi Freq
p2 <- ggplot(top_10_muni_freq, aes(y = reorder(municipio, n), x = n)) +
  geom_col(fill = blue_palette[2], width = 0.7) +
  geom_text(aes(label=n), hjust=-0.2, size=3, fontface="bold") +
  expand_limits(x = max(top_10_muni_freq$n) * 1.2) +
  labs(title = "Top 10 municipalities (frequency)", y = NULL, x = "Number of fires") +
  estil_tfm_horizontal()

# --- SEVERITY PLOTS (HORIZONTALS) ---

# P3: Província Area
p3 <- ggplot(area_provincia, aes(y = reorder(provincia, total_area), x = total_area)) +
  geom_col(fill = blue_palette[3], width = 0.7) +
  geom_text(aes(label=comma(round(total_area))), hjust=-0.2, size=3, fontface="bold") +
  scale_x_continuous(labels = comma) +
  expand_limits(x = max(area_provincia$total_area) * 1.2) +
  labs(title = "Burnt area by province", y = NULL, x = "Total area (ha)") +
  estil_tfm_horizontal()

# P4: Municipi Area
p4 <- ggplot(top_10_muni_area, aes(y = reorder(municipio, total_area), x = total_area)) +
  geom_col(fill = blue_palette[4], width = 0.7) +
  geom_text(aes(label=comma(round(total_area))), hjust=-0.2, size=3, fontface="bold") +
  scale_x_continuous(labels = comma) +
  expand_limits(x = max(top_10_muni_area$total_area) * 1.2) +
  labs(title = "Top 10 municipalities (burnt area)", y = NULL, x = "Total area (ha)") +
  estil_tfm_horizontal()

# 5. Assemblea final (2 columnes, 2 files)
final_plot <- (p1 + p2) / (p3 + p4) + 
  plot_annotation(
    title = 'Spatial impact analysis: frequency vs. severity',
    subtitle = 'Historical distribution across Catalonia (1998-2022)',
    theme = theme(
      plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "grey40")
    )
  )

print(final_plot)
```

### Key insights from the spatial analysis

The comparison between frequency and severity reveals a clear decoupling of fire patterns in Catalonia:

-   **The "Urban-Wildland Interface" (WUI) paradox**: Barcelona leads both in frequency (5,821 fires) and total area (35,430 ha), but at the municipal level, the pattern changes. Terrassa is the municipality with the highest number of ignitions (210), likely due to its large perimeter in contact with forest areas (high human activity). However, it does not even appear in the "Top 10" of burnt area. This suggests that while ignitions are frequent near large cities, suppression is usually very fast and effective.

-   **High-intensity hotspots**: Contrastingly, Aguilar de Segarra leads the severity ranking with 12,515 ha burnt, despite not being a leader in frequency. This indicates that this area is prone to Large Wildfires, where a single event can cause more ecological damage than hundreds of small ignitions in the metropolitan area.

-   **Strategic border areas**: La Jonquera stands out as a critical point. It ranks high in both frequency (187 fires) and burnt area (8,787 ha).

-   **Regional efficiency**: While Lleida has the lowest frequency and total burnt area among the four provinces, its municipalities like Artesa de Segre or Agramunt appear in the top severity list.

# 4. Chapter 3: Land Cover and ecosystem vulnerability

```{r}
# 1. Preparació de dades 
land_cover_summary <- wfc_final2 %>%
  group_by(land_cover_type) %>%
  summarise(
    frequency = n(),
    total_area = sum(superficie_total_forestal, na.rm = TRUE)
  ) %>%
  filter(land_cover_type != "No data")

top_10_lc_freq <- land_cover_summary %>% slice_max(frequency, n = 10)
top_10_lc_area <- land_cover_summary %>% slice_max(total_area, n = 10)

# Paleta de colors verds
tree_color <- "#2d6a4f"
trunk_color <- "#606c38"

# 2. Funció d'estil per a format HORITZONTAL
estil_forestal_horiz <- function() {
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 12, color = "#1b4332"),
    axis.text.y = element_text(size = 9), 
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(5, 20, 5, 5) 
  )
}

# --- GRÀFICS HORITZONTALS ---

# P5: Fire frequency by land cover type
p5 <- ggplot(top_10_lc_freq, aes(y = reorder(land_cover_type, frequency), x = frequency)) +
  geom_segment(aes(yend = reorder(land_cover_type, frequency), xend = 0), 
               color = trunk_color, size = 1.2) +
  geom_point(size = 7, color = tree_color) +
  geom_text(label = "🌲", size = 3, color = "white") +
  geom_text(aes(label = frequency), hjust = -0.6, size = 3.5, fontface = "bold", color = "#1b4332") +
  expand_limits(x = max(top_10_lc_freq$frequency) * 1.15) +
  labs(title = "Fire frequency by land cover type", y = NULL, x = "Number of fires") +
  estil_forestal_horiz()

# P6: Total burnt area by land cover type
p6 <- ggplot(top_10_lc_area, aes(y = reorder(land_cover_type, total_area), x = total_area)) +
  geom_segment(aes(yend = reorder(land_cover_type, total_area), xend = 0), 
               color = trunk_color, size = 1.2) +
  geom_point(size = 7, color = tree_color) +
  geom_text(label = "🌳", size = 3, color = "white") +
  # Etiqueta amb format de milers
  geom_text(aes(label = comma(round(total_area))), hjust = -0.4, size = 3.5, fontface = "bold", color = "#1b4332") +
  scale_x_continuous(labels = comma) +
  expand_limits(x = max(top_10_lc_area$total_area) * 1.2) +
  labs(title = "Total burnt area by land cover type", y = NULL, x = "Total area (ha)") +
  estil_forestal_horiz()

# 3. Unió final (un sota l'altre)
land_cover_tree_plot <- p5 / p6 + 
  plot_annotation(
    title = "Ecosystem Impact: Land cover analysis",
    subtitle = "Vegetation and soil types visualized as forest density (1998-2022)",
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5, face = "bold", color = "#1b4332"))
  )

print(land_cover_tree_plot)
```

The analysis of wildfire impact across different land cover types reveals a critical duality between ignition frequency and landscape severity (total burnt area):

-   **Vulnerability of coniferous forests**: Dense coniferous forests exhibit the **highest fire frequency** with **3,250 recorded events**. This suggests a high susceptibility to ignition, likely due to the high flammability of resinous species.

-   **The herbaceous crop paradox**: While ranking third in frequency, Herbaceous crops represent the largest total burnt area, **exceeding 25,500 hectares**. This indicates that although fires may start less frequently in agricultural zones compared to forests, they tend to propagate much faster across open fields, affecting vast extensions in shorter timeframes.

-   **Shrublands as a constant threat**: Shrublands maintain a high-pressure profile in both metrics, ranking second in frequency (2,745) and second in burnt area (22,276 ha). These areas often act as "ecological bridges," allowing low-intensity fires to transition from agricultural fringes into dense forest cores.

-   **Relative resilience of deciduous forests**: In contrast, dense deciduous broadleaf forests show significantly lower total burnt areas (7,661 ha). This may be attributed to higher fuel moisture retention and the lower flammability of broadleaf litter compared to needle-leaf species.

-   **Urban and infrastructure pressure**: The appearance of the Road network and Urban expansion in the burnt area rankings highlights the increasing risk at the Wildland-Urban Interface (WUI), where human activity acts as a primary driver of landscape transformation through fire.

# 5. Chapter 4: Analysis of ignition sources and drivers

```{r}
# 1. Preparació de dades de causes (Top 15)
causes_data <- wfc_final2 %>%
  group_by(causa) %>% 
  summarise(
    frequency = n(),
    total_area = sum(superficie_total_forestal, na.rm = TRUE)
  ) %>%
  filter(!is.na(causa) & causa != "No data")

# Seleccionem les 15 primeres per cada mètrica
top_15_causes_freq <- causes_data %>% slice_max(frequency, n = 10)
top_15_causes_area <- causes_data %>% slice_max(total_area, n = 10)

# 2. Paleta de colors "Fire" (taronges i vermells)
cause_palette <- c("#feb24c", "#fd8d3c", "#f03b20", "#bd0026")

# Funció d'estil optimitzada per a format horitzontal
estil_causes_tfm <- function() {
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 12, color = "#800026"),
    axis.text.y = element_text(size = 9), 
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    plot.margin = margin(5, 20, 5, 5)
  )
}

# --- GRÀFICS HORITZONTALS ---

# P7: Top 15 Causes by Frequency
p7 <- ggplot(top_15_causes_freq, aes(y = reorder(causa, frequency), x = frequency)) +
  geom_col(fill = cause_palette[2], width = 0.7) +
  geom_text(aes(label = frequency), hjust = -0.2, size = 3, fontface = "bold") +
  expand_limits(x = max(top_15_causes_freq$frequency) * 1.2) +
  labs(title = "Top 15 fire causes by frequency", y = NULL, x = "Number of fires") +
  estil_causes_tfm()

# P8: Top 15 Causes by Burnt Area
p8 <- ggplot(top_15_causes_area, aes(y = reorder(causa, total_area), x = total_area)) +
  geom_col(fill = cause_palette[4], width = 0.7) +
  geom_text(aes(label = scales::comma(round(total_area))), hjust = -0.2, size = 3, fontface = "bold") +
  scale_x_continuous(labels = scales::comma) +
  expand_limits(x = max(top_15_causes_area$total_area) * 1.25) +
  labs(title = "Top 15 fire causes by burnt area", y = NULL, x = "Total area (ha)") +
  estil_causes_tfm()

# 3. Assemblea final (un sota l'altre)
causes_plot_final <- p7 / p8 + 
  plot_annotation(
    title = "Analysis of ignition sources",
    subtitle = "Top 15 most frequent and impactful causes (1998-2022)",
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5, face = "bold", color = "#800026"))
  )

print(causes_plot_final)
```

The investigation into the origins of wildfires in Catalonia (1998-2022) reveals a landscape dominated by human activity, where the intent and negligence of individuals outweigh natural causes in both frequency and severity.

-   **The critical impact of arson**: The category "[400] Intencionado" is the primary driver of the fire problem in Catalonia. It not only records the highest frequency with **3,548 ignitions** but also results in the largest ecological footprint, accounting for **30,435 hectares burnt**.

-   **Negligence and infrastructure**: While intentional fires lead in severity, a significant portion of the total area is lost due to technical and human failures. "[320] Otras causas por líneas eléctricas" represents a major risk factor: although it is not in the top three for frequency, it ranks second in burnt area (**18,736 ha**). This suggests that utility-related fires often occur in remote or forested areas where detection is slower and suppression more complex.

-   **Smoking and public awareness**: The category "[250] Fumadores" remains a persistent threat, appearing prominently in both charts. With over **800 fires** and **14,255 hectares** destroyed, it underscores the ongoing need for public awareness campaigns regarding the disposal of cigarette butts, especially near road networks and forest interfaces.

-   **Natural vs. anthropogenic origins**: Lightning ("[100] Rayo") is the most frequent natural cause (**1,477 fires**). However, its impact on the total burnt area (**5,029 ha**) is considerably lower than that of intentional or negligent human acts. This confirms that wildfires in Catalonia are, fundamentally, a human-driven phenomenon.

-   **The challenge of uncertainty**: The high volume of "[500] Desconocida" cases (**1,809 ignitions**) points toward the difficulty of post-fire forensic investigation. Reducing this statistical "noise" is essential for improving targeted prevention measures.
