---
title: "Unveiling wildfire patterns in Catalonia (1998-2022)"
author: "Gemma Bargalló Solé"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---
```{r}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r warning=FALSE, message=FALSE}
# 1. packages import
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(nanoparquet)
library(here)
library(scales)
library(naniar) 
library(skimr) 
library(patchwork)
library(VIM)
library(gt)
library(purrr)
library(forcats)
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
library(leaflet)
library(dplyr)
library(crosstalk)
library(shiny)
library(shinydashboard)  
library(plotly)
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)
library(ranger)

# 2. Data Import
wfc_final2 <- read_csv(here("Data", "wfc_final2.csv"))

```

# Introduction: The fire challenge in Catalonia

Wildfires are one of the most significant environmental threats in Catalonia. As climate change intensifies and traditional land management fades, the Mediterranean landscape becomes increasingly vulnerable.

![Source: [El Periódico (2021). Wildfire risk factors in Catalonia.](https://www.elperiodico.cat/ca/successos/20210709/tres-claus-risc-incendis-forestals-11897390)](images/incendis%20catalunya.jpg){width="100%"}

## The objectives of this visualization:

-   **Identify spatio-temporal patterns**: To understand not just when it burns, but where the recurrence is highest.

-   **Correlate causes and landscape**: To see how human activity and land cover types interact to create risk.

-   **Bridge descriptive and predictive analysis**: To provide a visual foundation that justifies the variables used in our Random Forest predictive model.

"*Data is the first step toward effective prevention. By visualizing the past 25 years, we can better prepare for the next season.*"

```{r}
# In this first section, we can show a summary "Value Box" 
# to show the total magnitude of the dataset.
total_fires <- nrow(wfc_final2)
total_area <- sum(wfc_final2$superficie_total_forestal, na.rm = TRUE)

# Custom display using valueBoxes
fluidRow(
  valueBox(total_fires, subtitle = "Total Fires Recorded", icon = icon("fire"), color = "red"),
  valueBox(round(total_area / 1000, 1), subtitle = "Total Thousand Ha Burnt", icon = icon("tree"), color = "yellow")
)
```

# Chapter 1: The pulse of time (temporal analysis)

### Understanding 25 years of fire history {data-commentary-width="350"}

```{r warning=FALSE, message=FALSE, fig.height=7, fig.width=10}
# 2. Data Preparation
# 2.1. Yearly Data: Convert year to numeric to avoid factor errors in scales
yearly_counts <- wfc_final2 %>%
  count(year) %>%
  mutate(year = as.numeric(as.character(year)))

# 2.2. Monthly Data: Define chronological order for the factor
wfc_final2$month_detected <- factor(wfc_final2$month_detected, 
                                    levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

monthly_counts <- wfc_final2 %>%
  count(month_detected)

# 3. Create Plots

# Plot A: Yearly Evolution with Trend Line
p_year <- ggplot(yearly_counts, aes(x = year, y = n)) +
  # Add the trend line (Linear Regression)
  geom_smooth(method = "lm", color = "steelblue", linetype = "dashed", size = 0.8, se = FALSE) +
  # Keep the actual data points and lines
  geom_line(color = "darkred", size = 1) +
  geom_point(color = "darkred", size = 2) +
  # Force X-axis to show every single year vertically
  scale_x_continuous(breaks = seq(min(yearly_counts$year), max(yearly_counts$year), by = 1)) +
  theme_minimal() +
  labs(title = "Annual fire frequency evolution", 
       subtitle = "1998 - 2022 Historical Series with Linear Trend",
       x = "Year", 
       y = "Number of Fires") +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1) # Vertical labels
  )

# Plot B: Monthly Seasonality (Bar Chart)
p_month <- ggplot(monthly_counts, aes(x = month_detected, y = n, group = 1)) +
  geom_bar(stat = "identity", fill = "indianred3") +
  theme_minimal() +
  labs(title = "Monthly fire seasonality", 
       subtitle = "Aggregated frequency (1998-2022)",
       x = "Month", 
       y = "Number of fires") +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# 4. Final Layout Assembly
# Side-by-side distribution using patchwork
final_temporal_plot <- p_year + p_month + 
  plot_annotation(
    title = 'Temporal analysis of wildfires in Catalonia',
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5, face = "bold"))
  )

# 5. Display the result
print(final_temporal_plot)
```

To understand the current risk, we must look at the historical series (1998-2022). As shown in the Annual Wildfire Summary table and the temporal plots:

-   **Interannual variability**: Wildfires in Catalonia are episodic. The table highlights that years like 1998 and 2012 are not just frequency peaks but true ecological disasters, with 20,992 ha and 15,026 ha burnt respectively. These represent critical thresholds of drought and heat.

-   **Frequency vs. severity:** By comparing the line chart with the table, we notice an interesting pattern: a high number of fires (like in 2005 with 894 ignitions) doesn't always correlate with the largest burnt area (5,495 ha), suggesting that fire suppression efficiency varies depending on meteorological synoptics.

-   **The seasonal double peak**: While the bar chart shows summer is the main threat, the secondary peak in March highlights the danger of agricultural winter fires, a factor to be further analyzed in the "Causes" section.

-   **A slow descent**: The blue trend line in the plot, supported by the lower figures in the last decade of the table (e.g., 2018 or 2020), reflects a slight decrease in ignitions over time. This trend suggests that prevention measures are mitigating the increasing climatic risk.

### Detailed annual data

```{r}
# 1. Prepare the data
table_data <- wfc_final2 %>%
  group_by(year) %>%
  summarise(
    Total_Fires = n(),
    Total_Area_Ha = sum(superficie_total_forestal, na.rm = TRUE)
  ) %>%
  arrange(year)

# 2. Create the corrected gt table
fire_table <- table_data %>%
  gt() %>%
  tab_header(
    title = md("**Annual wildfire summary**"),
    subtitle = "Frequency and Burnt Surface (1998-2022)"
  ) %>%
  cols_label(
    year = "Year",
    Total_Fires = "Number of Fires",
    Total_Area_Ha = "Total Area (ha)"
  ) %>%
  fmt_number(
    columns = c(Total_Fires, Total_Area_Ha),
    decimals = 1,
    use_seps = TRUE
  ) %>%
  # Updated styling method
  tab_style(
    style = cell_fill(color = "#F9F9F9"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.width = pct(100),
    column_labels.font.weight = "bold"
  )

fire_table %>%
  tab_options(
    container.height = px(400), 
    container.overflow.y = TRUE 
  )

```

# Chapter 2: The core of the story (interactive exploration)

```{r}
wfc_final2 <- wfc_final2 %>%
  mutate(
    # Fem el canvi basant-nos en la fila (row_number())
    municipio = case_when(
      row_number() == 778 ~ "AGUILAR DE SEGARRA",
      row_number() == 779 ~ "CARDONA",
      row_number() == 2096 ~ "OLIVELLA",
      row_number() == 2386 ~ "LA SÈNIA",
      row_number() == 2919 ~ "VALLCLARA",
      row_number() == 3348 ~ "MONTMANEU",
      row_number() == 3516 ~ "TORDERA",
      row_number() == 3524 ~ "SENAN",
      row_number() == 4740 ~ "SERÒS",
      row_number() == 4793 ~ "MARGALEF",
      row_number() == 4797 ~ "CARDONA",
      row_number() == 6068 ~ "SANT SALVADOR DE TORROELLA",
      row_number() == 7287 ~ "CASTELLFOLLIT DE RIUBREGÓS",
      row_number() == 8810 ~ "EL PLA DE MANLLEU",
      TRUE ~ municipio 
    ),
    
    provincia = case_when(
      row_number() == 778 ~ "BARCELONA",
      row_number() == 779 ~ "BARCELONA",
      row_number() == 2096 ~ "BARCELONA",
      row_number() == 2386 ~ "TARRAGONA",
      row_number() == 2919 ~ "TARRAGONA",
      row_number() == 3348 ~ "BARCELONA",
      row_number() == 3516 ~ "BARCELONA",
      row_number() == 3524 ~ "TARRAGONA",
      row_number() == 4740 ~ "LLEIDA",
      row_number() == 4793 ~ "TARRAGONA",
      row_number() == 4797 ~ "BARCELONA",
      row_number() == 6068 ~ "BARCELONA",
      row_number() == 7287 ~ "BARCELONA",
      row_number() == 8810 ~ "TARRAGONA",
      TRUE ~ provincia
    ),
    
    county_clean = case_when(
      row_number() == 778 ~ "BAGES",
      row_number() == 779 ~ "BAGES",
      row_number() == 2096 ~ "GARRAF",
      row_number() == 2386 ~ "MONTSIÀ",
      row_number() == 2919 ~ "CONCA DE BARBERÀ",
      row_number() == 3348 ~ "ANOIA",
      row_number() == 3516 ~ "MARESME",
      row_number() == 3524 ~ "CONCA DE BARBERÀ",
      row_number() == 4740 ~ "SEGRIÀ",
      row_number() == 4793 ~ "PRIORAT",
      row_number() == 4797 ~ "BAGES",
      row_number() == 6068 ~ "BAGES",
      row_number() == 7287 ~ "ANOIA",
      row_number() == 8810 ~ "ALT CAMP",
      TRUE ~ county_clean
    )
  )
```

```{r}
# Creem el dataframe de referència (diccionari)
land_cover_lookup <- data.frame(
  land_cover_id = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 
                    19, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 30, 31, 32, 33, 
                    34, 35, 36, 37, 38, 39, 40, 41, 0),
  land_cover_type = c(
    "Herbaceous crops", "Market gardens, nurseries and greenhouse crops", 
    "Vineyards", "Olive groves", "Other woody crops", "Crops in transformation", 
    "Dense coniferous forests", "Dense deciduous broadleaf forests", 
    "Dense sclerophyllous and laurel forests", "Shrubland", "Open coniferous forests", 
    "Open deciduous broadleaf forests", "Open sclerophyllous and laurel forests", 
    "Grasslands", "Riparian forest", "Bare forest soil", "Burnt areas", 
    "Rocky areas and scree", "Beaches", "Wetlands", "Urban core", 
    "Urban expansion", "Low-density urban areas", "Isolated buildings in rural areas", 
    "Isolated residential areas", "Green urban areas", "Industrial, commercial and/or service areas", 
    "Sports and leisure areas", "Mining extraction areas and/or landfills", 
    "Areas under transformation", "Road network", "Bare urban soil", 
    "Airport areas", "Railway network", "Port areas", "Reservoirs", 
    "Lakes and lagoons", "Watercourses", "Ponds", "Artificial canals", "Sea", "No data"
  )
)
# Comprovem que land_cover_id sigui numeric en ambdós costats
wfc_final2$land_cover_id <- as.numeric(as.character(wfc_final2$land_cover_id))

# Afegim la descripció
wfc_final2 <- wfc_final2 %>%
  left_join(land_cover_lookup, by = "land_cover_id")
```

```{r}
# MAP
# 1. Data Preparation for Leaflet
wfc_map_data <- wfc_final2 %>%
  mutate(
    # Year MUST be numeric for the slider to work correctly
    year_num = as.numeric(as.character(year)),
    
    # Create the customized popup text in English
    popup_text = paste0(
      "<div style='font-family: Arial; font-size: 12px;'>",
      "<b style='color: #8B0000;'>Municipality:</b> ", municipio, "<br>",
      "<strong>Year:</strong> ", year, "<br>",
      "<strong>Detected:</strong> ", date_detected, "<br>",
      "<strong>Extinguished:</strong> ", date_extinguished, "<br>",
      "<strong>Burnt Surface:</strong> ", superficie_total_forestal, " ha<br>",
      "<strong>Land Cover Type:</strong> ", land_cover_type, "<br>",
      "<strong>Cause:</strong> ", causa, 
      "</div>"
    )
  ) %>%
  # Seleccionem les variables necessàries i traiem el land_cover_id
  select(longitude, latitude, year_num, popup_text, superficie_total_forestal, 
         municipio, date_detected, date_extinguished, land_cover_type)

# 2. Create SharedData object
sd <- SharedData$new(wfc_map_data)

# 3. Interactive Layout with Slider
bscols(
  widths = c(12),
  # Year Slider
  filter_slider(
    id = "year_slider", 
    label = "Slide to navigate through years (1998-2022):", 
    sharedData = sd, 
    column = ~year_num,
    step = 1,          # Move year by year
    ticks = TRUE,      # Show year marks
    sep = "",          # Avoid commas in years (e.g., 2,022 -> 2022)
    width = "100%"
  ),
  # Leaflet Map
  leaflet(sd, width = "100%", height = 700) %>%
    addTiles() %>% 
    addProviderTiles(providers$CartoDB.Positron) %>%
    addCircleMarkers(
      lng = ~longitude, 
      lat = ~latitude,
      radius = ~sqrt(superficie_total_forestal) + 3, 
      color = "#E44D2E",
      stroke = TRUE,
      weight = 1,
      fillOpacity = 0.7,
      popup = ~popup_text,
      label = ~paste(municipio, "-", superficie_total_forestal, "ha")
    )
)
```

### Where does Catalonia burn?

This interactive map bridges the gap between raw data and territorial management:

-   **Regional pressure**: You can visually confirm the high density of events in the Barcelona Metropolitan Area and the Empordà.

-   **Magnitude vs. frequency**: Use the slider to see how certain years were dominated by many small fires, while others were defined by a few Large Forest Fires.

-   **Spatial patterns**: This visualization justifies why Latitude and Longitude became the most important variables in our predictive model.

In addition to the interactive map, we have generated these bar charts to summarize the provinces and municipalities with the highest fire frequency and the most significant burnt area. This dual perspective allows us to distinguish between areas of high human or natural activity (frequency) and areas of high ecological vulnerability (burnt area).

```{r}
# 1. Preparació de dades (Top 10)
top_10_muni_freq <- wfc_final2 %>%
  count(municipio) %>%
  slice_max(n, n = 10)

top_10_muni_area <- wfc_final2 %>%
  group_by(municipio) %>%
  summarise(total_area = sum(superficie_total_forestal, na.rm = TRUE)) %>%
  slice_max(total_area, n = 10)

area_provincia <- wfc_final2 %>%
  group_by(provincia) %>%
  summarise(total_area = sum(superficie_total_forestal, na.rm = TRUE))

# Paleta de blaus
blue_palette <- c("#08306b", "#08519c", "#2171b5", "#4292c6")

# Funció d'estil optimitzada per a gràfics horitzontals
estil_tfm_horizontal <- function() {
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 11, color = "#2c3e50"),
    axis.text.y = element_text(size = 9), # Noms de municipis més clars
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank(), # Treiem graella horitzontal
    panel.grid.minor = element_blank(),
    plot.margin = margin(5, 15, 5, 5)
  )
}

# --- FREQUENCY PLOTS (HORIZONTALS) ---

# P1: Província Freq
p1 <- ggplot(wfc_final2, aes(y = reorder(provincia, provincia, function(x) length(x)), x = ..count..)) +
  geom_bar(fill = blue_palette[1], width = 0.7) +
  geom_text(stat='count', aes(label=..count..), hjust=-0.2, size=3, fontface="bold") +
  expand_limits(x = count(wfc_final2, provincia) %>% pull(n) %>% max() * 1.2) +
  labs(title = "Fire frequency by province", y = NULL, x = "Number of fires") +
  estil_tfm_horizontal()

# P2: Municipi Freq
p2 <- ggplot(top_10_muni_freq, aes(y = reorder(municipio, n), x = n)) +
  geom_col(fill = blue_palette[2], width = 0.7) +
  geom_text(aes(label=n), hjust=-0.2, size=3, fontface="bold") +
  expand_limits(x = max(top_10_muni_freq$n) * 1.2) +
  labs(title = "Top 10 municipalities (frequency)", y = NULL, x = "Number of fires") +
  estil_tfm_horizontal()

# --- SEVERITY PLOTS (HORIZONTALS) ---

# P3: Província Area
p3 <- ggplot(area_provincia, aes(y = reorder(provincia, total_area), x = total_area)) +
  geom_col(fill = blue_palette[3], width = 0.7) +
  geom_text(aes(label=comma(round(total_area))), hjust=-0.2, size=3, fontface="bold") +
  scale_x_continuous(labels = comma) +
  expand_limits(x = max(area_provincia$total_area) * 1.2) +
  labs(title = "Burnt area by province", y = NULL, x = "Total area (ha)") +
  estil_tfm_horizontal()

# P4: Municipi Area
p4 <- ggplot(top_10_muni_area, aes(y = reorder(municipio, total_area), x = total_area)) +
  geom_col(fill = blue_palette[4], width = 0.7) +
  geom_text(aes(label=comma(round(total_area))), hjust=-0.2, size=3, fontface="bold") +
  scale_x_continuous(labels = comma) +
  expand_limits(x = max(top_10_muni_area$total_area) * 1.2) +
  labs(title = "Top 10 municipalities (burnt area)", y = NULL, x = "Total area (ha)") +
  estil_tfm_horizontal()

# 5. Assemblea final (2 columnes, 2 files)
final_plot <- (p1 + p2) / (p3 + p4) + 
  plot_annotation(
    title = 'Spatial impact analysis: frequency vs. severity',
    subtitle = 'Historical distribution across Catalonia (1998-2022)',
    theme = theme(
      plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "grey40")
    )
  )

print(final_plot)
```

### Key insights from the spatial analysis

The comparison between frequency and severity reveals a clear decoupling of fire patterns in Catalonia:

-   **The "Urban-Wildland Interface" (WUI) paradox**: Barcelona leads both in frequency (5,821 fires) and total area (35,430 ha), but at the municipal level, the pattern changes. Terrassa is the municipality with the highest number of ignitions (210), likely due to its large perimeter in contact with forest areas (high human activity). However, it does not even appear in the "Top 10" of burnt area. This suggests that while ignitions are frequent near large cities, suppression is usually very fast and effective.

-   **High-intensity hotspots**: Contrastingly, Aguilar de Segarra leads the severity ranking with 12,515 ha burnt, despite not being a leader in frequency. This indicates that this area is prone to Large Wildfires, where a single event can cause more ecological damage than hundreds of small ignitions in the metropolitan area.

-   **Strategic border areas**: La Jonquera stands out as a critical point. It ranks high in both frequency (187 fires) and burnt area (8,787 ha).

-   **Regional efficiency**: While Lleida has the lowest frequency and total burnt area among the four provinces, its municipalities like Artesa de Segre or Agramunt appear in the top severity list.

# Chapter 3: The meteorological dimension

```{r warning=FALSE, message=FALSE}
# Gràfic de densitat: On es concentren els incendis segons Clima?
ggplot(wfc_final2, aes(x = tx, y = ppt)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", color = "white") +
  geom_point(aes(size = superficie_total_forestal), alpha = 0.2, color = "orange") +
  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_size_continuous(range = c(1, 10), name = "Area (ha)") +
  theme_minimal() +
  labs(
    title = "Meteorological fingerprint of wildfires",
    subtitle = "Interaction between Max Temperature (TX) and Precipitation (PPT)",
    x = "Max Temperature (°C)",
    y = "Precipitation (mm)"
  )
```

### TX and PPT Interaction

The scatter plot and density analysis presented in the previous plot illustrate the climatic envelope in which wildfires occur in Catalonia. This "fingerprint" reveals a critical dependence on specific atmospheric conditions for fire spread and intensification.

-   **The zero-precipitation constraint**: A striking pattern is the extreme concentration of ignition events along the 0 mm precipitation line. The density heatmap (indicated by the lighter yellow tones) confirms that the vast majority of fires occur under conditions of absolute lack of immediate rainfall. This confirms that while a fire can start at various temperatures, it rarely prospers or even occurs if there is significant moisture from recent precipitation.

-   **The thermal threshold for severity**: When observing the size of the points (representing **area in ha**), there is a clear shift as we move toward the right of the X-axis. Although fires are frequent between 20°C and 30°C, the **Large Wildfires**---represented by the largest bubbles---are almost exclusively clustered between **25°C and 40°C**. This thermal window represents the "danger zone" where suppression efforts face the greatest challenges due to fuel volatility.

-   **Atmospheric "Dry windows"**: There are occasional ignitions at higher precipitation levels (up to 60-70 mm), but these remain isolated, small-scale events (dots with almost no radius). This suggests that even if an ignition occurs during a rainy period, the lack of fuel continuity or the high fuel moisture content prevents the fire from gaining significant size.

-   **Operational implications**: The "Meteorological Fingerprint" suggests that the risk is not linear but threshold-based. Once precipitation reaches 0 mm and temperature crosses the 30°C mark, the landscape enters a state of high receptivity where any anthropogenic ignition (as we will analyze in Chapter 5) has a high probability of becoming a high-impact event.

```{r warning=FALSE, message=FALSE}
# Evolució de la TX mitjana per any
climate_evolution <- wfc_final2 %>%
  group_by(year) %>%
  summarise(avg_tx = mean(tx, na.rm = TRUE))

ggplot(climate_evolution, aes(x = year, y = avg_tx)) +
  geom_line(color = "red", size = 1) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", linetype = "dashed", color = "darkred") +
  theme_minimal() +
  labs(
    title = "Thermal evolution of fire days (1998-2022)",
    subtitle = "Annual average of TX during ignition events",
    x = "Year",
    y = "Average Maximum Temperature (°C)"
  )
```

### Temporal trends: the warming of fire days (1998-2022)

The historical evolution of the average maximum temperature (**TX**) during ignition events, as illustrated in the previous plot, provides a direct signal of climate change's influence on the regional fire regime.

-   **Positive linear trend**: The dashed regression line shows a steady upward trend over the 25-year period. This indicates that, on average, wildfires in Catalonia are occurring under increasingly hotter conditions. Even if the slope appears gradual, in ecological terms, a consistent rise in the thermal baseline significantly reduces fuel moisture and expands the annual "fire window."

-   **Interannual volatility**: The red line shows high variability, with extreme thermal peaks such as the one recorded in **2003** (surpassing 30°C on average). This specific peak correlates with one of the most severe heatwaves in European history, which led to catastrophic wildfire seasons in the Mediterranean. Similar, though less extreme, peaks are visible in **2006**, **2009**, and more recently in the **2017-2020** period.

-   **A new thermal plateau**: It is noteworthy that since 2015, the annual average **TX** for fire days has rarely dropped below 26°C, whereas in the first decade of the series (1998-2008), values below 25°C were much more frequent. This suggests a "thermal hardening" of the environment, where the minimum conditions required for fire propagation are becoming more extreme.

-   **The confidence interval**: The grey shaded area (confidence interval) widening towards the end of the series reflects the increasing unpredictability and frequency of extreme events. While some years still show lower averages due to specific rainy summers, the overall trajectory is towards a hotter ignitions environment.

# Chapter 4: Land cover and ecosystem vulnerability

```{r}
# 1. Preparació de dades 
land_cover_summary <- wfc_final2 %>%
  group_by(land_cover_type) %>%
  summarise(
    frequency = n(),
    total_area = sum(superficie_total_forestal, na.rm = TRUE)
  ) %>%
  filter(land_cover_type != "No data")

top_10_lc_freq <- land_cover_summary %>% slice_max(frequency, n = 10)
top_10_lc_area <- land_cover_summary %>% slice_max(total_area, n = 10)

# Paleta de colors verds
tree_color <- "#2d6a4f"
trunk_color <- "#606c38"

# 2. Funció d'estil per a format HORITZONTAL
estil_forestal_horiz <- function() {
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 12, color = "#1b4332"),
    axis.text.y = element_text(size = 9), 
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(5, 20, 5, 5) 
  )
}

# --- GRÀFICS HORITZONTALS ---

# P5: Fire frequency by land cover type
p5 <- ggplot(top_10_lc_freq, aes(y = reorder(land_cover_type, frequency), x = frequency)) +
  geom_segment(aes(yend = reorder(land_cover_type, frequency), xend = 0), 
               color = trunk_color, size = 1.2) +
  geom_point(size = 7, color = tree_color) +
  geom_text(label = "🌲", size = 3, color = "white") +
  geom_text(aes(label = frequency), hjust = -0.6, size = 3.5, fontface = "bold", color = "#1b4332") +
  expand_limits(x = max(top_10_lc_freq$frequency) * 1.15) +
  labs(title = "Fire frequency by land cover type", y = NULL, x = "Number of fires") +
  estil_forestal_horiz()

# P6: Total burnt area by land cover type
p6 <- ggplot(top_10_lc_area, aes(y = reorder(land_cover_type, total_area), x = total_area)) +
  geom_segment(aes(yend = reorder(land_cover_type, total_area), xend = 0), 
               color = trunk_color, size = 1.2) +
  geom_point(size = 7, color = tree_color) +
  geom_text(label = "🌳", size = 3, color = "white") +
  # Etiqueta amb format de milers
  geom_text(aes(label = comma(round(total_area))), hjust = -0.4, size = 3.5, fontface = "bold", color = "#1b4332") +
  scale_x_continuous(labels = comma) +
  expand_limits(x = max(top_10_lc_area$total_area) * 1.2) +
  labs(title = "Total burnt area by land cover type", y = NULL, x = "Total area (ha)") +
  estil_forestal_horiz()

# 3. Unió final (un sota l'altre)
land_cover_tree_plot <- p5 / p6 + 
  plot_annotation(
    title = "Ecosystem Impact: Land cover analysis",
    subtitle = "Vegetation and soil types visualized as forest density (1998-2022)",
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5, face = "bold", color = "#1b4332"))
  )

print(land_cover_tree_plot)
```

The analysis of wildfire impact across different land cover types reveals a critical duality between ignition frequency and landscape severity (total burnt area):

-   **Vulnerability of coniferous forests**: Dense coniferous forests exhibit the **highest fire frequency** with **3,250 recorded events**. This suggests a high susceptibility to ignition, likely due to the high flammability of resinous species.

-   **The herbaceous crop paradox**: While ranking third in frequency, Herbaceous crops represent the largest total burnt area, **exceeding 25,500 hectares**. This indicates that although fires may start less frequently in agricultural zones compared to forests, they tend to propagate much faster across open fields, affecting vast extensions in shorter timeframes.

-   **Shrublands as a constant threat**: Shrublands maintain a high-pressure profile in both metrics, ranking second in frequency (2,745) and second in burnt area (22,276 ha). These areas often act as "ecological bridges," allowing low-intensity fires to transition from agricultural fringes into dense forest cores.

-   **Relative resilience of deciduous forests**: In contrast, dense deciduous broadleaf forests show significantly lower total burnt areas (7,661 ha). This may be attributed to higher fuel moisture retention and the lower flammability of broadleaf litter compared to needle-leaf species.

-   **Urban and infrastructure pressure**: The appearance of the Road network and Urban expansion in the burnt area rankings highlights the increasing risk at the Wildland-Urban Interface (WUI), where human activity acts as a primary driver of landscape transformation through fire.

# Chapter 5: Analysis of ignition sources and drivers

```{r}
# 1. Preparació de dades de causes (Top 15)
causes_data <- wfc_final2 %>%
  group_by(causa) %>% 
  summarise(
    frequency = n(),
    total_area = sum(superficie_total_forestal, na.rm = TRUE)
  ) %>%
  filter(!is.na(causa) & causa != "No data")

# Seleccionem les 15 primeres per cada mètrica
top_15_causes_freq <- causes_data %>% slice_max(frequency, n = 10)
top_15_causes_area <- causes_data %>% slice_max(total_area, n = 10)

# 2. Paleta de colors "Fire" (taronges i vermells)
cause_palette <- c("#feb24c", "#fd8d3c", "#f03b20", "#bd0026")

# Funció d'estil optimitzada per a format horitzontal
estil_causes_tfm <- function() {
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 12, color = "#800026"),
    axis.text.y = element_text(size = 9), 
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    plot.margin = margin(5, 20, 5, 5)
  )
}

# --- GRÀFICS HORITZONTALS ---

# P7: Top 15 Causes by Frequency
p7 <- ggplot(top_15_causes_freq, aes(y = reorder(causa, frequency), x = frequency)) +
  geom_col(fill = cause_palette[2], width = 0.7) +
  geom_text(aes(label = frequency), hjust = -0.2, size = 3, fontface = "bold") +
  expand_limits(x = max(top_15_causes_freq$frequency) * 1.2) +
  labs(title = "Top 15 fire causes by frequency", y = NULL, x = "Number of fires") +
  estil_causes_tfm()

# P8: Top 15 Causes by Burnt Area
p8 <- ggplot(top_15_causes_area, aes(y = reorder(causa, total_area), x = total_area)) +
  geom_col(fill = cause_palette[4], width = 0.7) +
  geom_text(aes(label = scales::comma(round(total_area))), hjust = -0.2, size = 3, fontface = "bold") +
  scale_x_continuous(labels = scales::comma) +
  expand_limits(x = max(top_15_causes_area$total_area) * 1.25) +
  labs(title = "Top 15 fire causes by burnt area", y = NULL, x = "Total area (ha)") +
  estil_causes_tfm()

# 3. Assemblea final (un sota l'altre)
causes_plot_final <- p7 / p8 + 
  plot_annotation(
    title = "Analysis of ignition sources",
    subtitle = "Top 15 most frequent and impactful causes (1998-2022)",
    theme = theme(plot.title = element_text(size = 16, hjust = 0.5, face = "bold", color = "#800026"))
  )

print(causes_plot_final)
```

The investigation into the origins of wildfires in Catalonia (1998-2022) reveals a landscape dominated by human activity, where the intent and negligence of individuals outweigh natural causes in both frequency and severity.

-   **The critical impact of arson**: The category "[400] Intencionado" is the primary driver of the fire problem in Catalonia. It not only records the highest frequency with **3,548 ignitions** but also results in the largest ecological footprint, accounting for **30,435 hectares burnt**.

-   **Negligence and infrastructure**: While intentional fires lead in severity, a significant portion of the total area is lost due to technical and human failures. "[320] Otras causas por líneas eléctricas" represents a major risk factor: although it is not in the top three for frequency, it ranks second in burnt area (**18,736 ha**). This suggests that utility-related fires often occur in remote or forested areas where detection is slower and suppression more complex.

-   **Smoking and public awareness**: The category "[250] Fumadores" remains a persistent threat, appearing prominently in both charts. With over **800 fires** and **14,255 hectares** destroyed, it underscores the ongoing need for public awareness campaigns regarding the disposal of cigarette butts, especially near road networks and forest interfaces.

-   **Natural vs. anthropogenic origins**: Lightning ("[100] Rayo") is the most frequent natural cause (**1,477 fires**). However, its impact on the total burnt area (**5,029 ha**) is considerably lower than that of intentional or negligent human acts. This confirms that wildfires in Catalonia are, fundamentally, a human-driven phenomenon.

-   **The challenge of uncertainty**: The high volume of "[500] Desconocida" cases (**1,809 ignitions**) points toward the difficulty of post-fire forensic investigation. Reducing this statistical "noise" is essential for improving targeted prevention measures.

### Behavioral drivers: A detailed analysis of arsonist motivations

```{r}
# 1. Preparació de dades amb cerca parcial de text
motivacions_data <- wfc_final2 %>%
  # Busquem qualsevol registre que contingui la paraula "Intencionado" sense importar majúscules
  filter(grepl("Intencionado", causa, ignore.case = TRUE)) %>% 
  group_by(motivacion) %>%
  summarise(
    frequency = n(),
    total_area = sum(superficie_total_forestal, na.rm = TRUE)
  ) %>%
  # Eliminem valors buits o NAs que solen embrutar el gràfic
  filter(!is.na(motivacion) & motivacion != "" & motivacion != "Desconeguda" & motivacion != "n/a") %>%
  slice_max(frequency, n = 10)

# d'aquesta manera veuràs com es diu exactament la categoria dels intencionats.

# 3. El gràfic
ggplot(motivacions_data, aes(y = reorder(motivacion, frequency), x = frequency)) +
  geom_col(fill = "#800026", width = 0.7) + 
  geom_text(aes(label = frequency), hjust = -0.2, size = 3.5, fontface = "bold", color = "#800026") +
  # Apliquem el salt de línia automàtic a l'eix Y
  scale_y_discrete(labels = function(x) str_wrap(x, width = 70)) +
  expand_limits(x = max(motivacions_data$frequency) * 1.15) +
  theme_minimal() +
  labs(
    title = "Analysis of arsonist motivations",
    subtitle = "Specific drivers within intentional fires",
    x = "Number of fires",
    y = NULL
  )
```

To refine the prevention strategies against intentional fires, it is essential to deconstruct the specific motivations behind the "[400] Intencionado" category. As shown in the previous plot, the breakdown of motivations provides a complex socio-behavioral map of wildfire origins in Catalonia:

-   **The challenge of unknown intent**: The most frequent category is **"[400] Motivación desconocida"** with **1,270 cases**. This high volume underscores the inherent difficulty in fire investigation; while the intentionality is proven by the presence of multiple ignition points or accelerants, the specific underlying grievance or objective often remains elusive.

-   **Vandalism and mental health**: A critical finding is the high incidence of **"[482] Vandalismo" (897 fires)** and **"[483] Enfermos mentales / Pirómanos" (761 fires)**. Together, these two categories represent a significant portion of the total intentional ignitions.

-   **Traditional and socio-economic conflicts**: Although less frequent than vandalism, motivations linked to land use remain persistent. These include **agricultural practices ([401])**, **livestock management ([402])**, and **hunting interests ([411])**. These categories reflect a historical friction between traditional rural activities and modern forest conservation regulations, where fire is still used as a prohibited tool for landscape management.

-   **Atypical drivers**: The data also captures marginal but significant behaviors, such as **"[464] Venganzas"** (Revenge) and, interestingly, **"[471] Incendios para distraer a la policía"**. These indicators highlight that wildfires are occasionally used as instruments of social conflict or as tactical diversions for other criminal activities.

# Chapter 6: The climate change and thermal extremes

```{r warning=FALSE, message=FALSE}
# 2. Afegim una columna de text personalitzada per al "tooltip" (el que es llegirà en passar el ratolí)
# Això farà que la interacció sigui molt més informativa.
wfc_final2 <- wfc_final2 %>%
  mutate(text_interactiu = paste0("Municipi: ", municipio, 
                                  "<br>Any: ", year,
                                  "<br>Temp. Máxima (TX): ", tx, "°C",
                                  "<br>Superfície: ", comma(superficie_total_forestal), " ha",
                                  "<br>Amplitud Tèrmica: ", round(thermal_amplitude, 1), "°C"))

# 3. Creem el ggplot assignant el text al paràmetre 'text'
p_clima <- ggplot(wfc_final2, aes(x = tx, y = superficie_total_forestal, text = text_interactiu)) +
  geom_point(aes(color = thermal_amplitude), alpha = 0.5, size = 2) +
  scale_y_log10(labels = scales::comma) + 
  scale_color_viridis_c(option = "inferno", name = "Thermal Amplitude") +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "The heat threshold: Max temperature vs. fire severity",
    x = "Maximum temperature (TX) at ignition (°C)",
    y = "Total burnt area (ha) [log10]"
  )

# 4. Convertim a gràfic interactiu
# L'argument tooltip="text" fa que només es mostri la nostra etiqueta personalitzada
ggplotly(p_clima, tooltip = "text")
```

In the Mediterranean basin, climate change acts as a "force multiplier." It does not necessarily increase the number of ignitions---which remain largely anthropogenic---but it drastically alters the environment's receptivity to fire. This chapter analyzes how thermal thresholds dictate the transition from routine fire events to uncontrollable Large Wildfires.

## Thermal thresholds and fire severity

The relationship between **Maximum Temperature (TX)** and the **total burnt area** provides a clear visualization of the "heat threshold".

-   **The 30°C inflection point**: Statistical density increases significantly when TX exceeds 30°C. While suppression forces are highly effective at lower temperatures, the "ceiling" of fire size rises exponentially once this thermal barrier is crossed.

-   **The extremes vs. the average**: Although the linear trend of fire size remains relatively stable across the temperature spectrum due to efficient initial attack, the most severe events (\>500 ha) are exclusively concentrated in the high-temperature range. This confirms that extreme heat creates a "window of opportunity" for fires to escape suppression.

## Thermal amplitude and fuel desiccation

The variable **`thermal_amplitude`** serves as a critical proxy for relative humidity and atmospheric stability.

-   **Atmospheric dryness**: High amplitude values (where $TX - TN > 20°C$) are strongly correlated with the most impactful wildfires in the 25-year series.

-   **Fire intensity**: Fires occurring under high thermal amplitude conditions exhibit higher flame lengths and faster rates of spread, as the low overnight recovery (**TN**) prevents the fuel from regaining moisture.

# Chapter 7: Predictive Modeling using Random Forest

## Model architecture and objective

In this stage, a Random Forest regressor was implemented to quantify the predictive power of environmental and anthropogenic variables over fire severity (Burnt Area). Random Forest was selected due to its robustness against outliers and its ability to capture non-linear relationships, which are prevalent in wildfire dynamics.

```{r}
# 1. Feature Selection and Formatting
# FINAL MODEL DATASET SELECTION
wfc_model_final <- wfc_final2 %>%
  select(
    # Target (Change this based on your goal)
    superficie_total_forestal, 
    
    # Location
    county_clean, altitude_z, latitude, longitude,
    
    # Meteorology
    tx, ppt, thermal_amplitude,
    
    # Time
    month_detected, hour_detected,year,
    
    # Land Cover & Risk
    land_cover_id, afecto_zonas_interfaz_urbano_forestal, 
    afecto_espacio_protegido, afecto_zar,
    
    # Context
    causa
  ) %>%
  # Convert to factors for Random Forest
  mutate(across(where(is.character), as.factor),
         land_cover_id = as.factor(land_cover_id),
         month_detected = as.factor(month_detected),
         # If hour_detected is numeric, we might keep it as is or factorize it
         hour_detected = as.numeric(hour_detected)) %>%
  drop_na()

# Apply Log1p transformation (log(x + 1)) to handle 0 values
wfc_model_final <- wfc_model_final %>%
  mutate(log_surface = log1p(superficie_total_forestal))

# Set seed for reproducibility
set.seed(123)

# Create the partition based on the target variable
train_index <- createDataPartition(wfc_model_final$log_surface, p = 0.8, list = FALSE)

# Generate sets
train_set <- wfc_model_final[train_index, ]
test_set  <- wfc_model_final[-train_index, ]


# Train the Random Forest model
# We predict 'log_surface' using all other columns in 'train_set'
rf_model <- ranger(
  formula         = log_surface ~ ., 
  data            = train_set %>% select(-superficie_total_forestal), # Exclude the original non-log surface
  num.trees       = 500,
  importance      = "permutation", # Important to analyze variable impact later
  seed            = 123
)

# Get importance
importance_values <- importance(rf_model)
importance_df <- data.frame(
  Variable = names(importance_values),
  Importance = importance_values
) %>% arrange(desc(Importance))

# Plot importance
ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Variable Importance in Wildfire Surface Prediction",
       x = "Predictors", y = "Importance (Permutation)")

# Predict on test set
predictions <- predict(rf_model, data = test_set)$predictions

# Calculate Performance Metrics (RMSE and R2)
postResample(pred = predictions, obs = test_set$log_surface)
```

## Model setup and performance metrics

To quantify the drivers of fire severity, a **Random Forest** regression model was developed using the `ranger` engine. The target variable, `superficie_total_forestal`, was transformed using a **log1p** function to normalize its highly skewed distribution and handle the high frequency of small-scale events.

The model evaluation on the test set yielded the following metrics:

-   **RMSE**: 0.692

-   **R-squared (**$R^2$): 0.151

-   **MAE**: 0.377

While an $R^2$ of **15.1%** might appear low in other disciplines, in wildfire science, it reflects the inherent complexity and "noise" of fire behavior. Predicting the final size of a fire at the moment of ignition is a major challenge, as it depends on dynamic factors not fully captured in static datasets, such as suppression efficiency, wind shifts, and fuel connectivity.

## Drivers of severity: variable importance analysis

The primary value of the model lies in its **Permutation Importance**, which identifies which factors most significantly influence the final burnt area:

-   **Geospatial predominance**: **Longitude** and **Latitude** are the top predictors. This suggests a strong spatial component in Catalonia, where specific geographic corridors are more prone to large-scale events due to persistent wind patterns or topographical funnels.

-   **The "WUI" and territorial context**: The high importance of **`county_clean`** and **`afecto_zonas_interfaz_urbano_forestal`** (WUI) confirms that the territorial mosaic and proximity to human settlements are critical. Fires near urban interfaces are often more complex to manage, as resources must be split between forest suppression and structural protection.

-   **Topography and climate**: **`altitude_z`** and **`tx`** (Maximum Temperature) appear as key environmental drivers. Higher temperatures lower the fuel's ignition threshold, while altitude influences both vegetation type and accessibility for ground crews.

-   **Anthropogenic drivers**: The variable **`causa`** ranks significantly high, reinforcing the idea that intentionality or specific types of negligence (as seen in Chapter 3) are not just frequency issues but indicators of potentially more severe fires.

## Discussion and conclusions

The Random Forest results validate that fire severity in Catalonia is a multi-causal phenomenon. The model successfully identifies **location and thermal stress (TX)** as the primary "engines" of fire size.

Interestingly, the low importance of **`ppt`** (Precipitation) suggests that, for the studied series, the immediate temperature and location are more decisive at the moment of ignition than recent rainfall. These insights are vital for forest management, indicating that resources should be strategically pre-positioned based on **geographic risk zones** and **thermal peaks** rather than relying solely on general seasonal trends.
