---
title: "EDA"
author: "Gemma Bargalló Solé"
date: "2026-04-20"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      position: left
    toc_depth: 5   
  word_document: default
  pdf_document: default
---

```{r packages import, warning=FALSE, message=FALSE}
library(tidyverse)
library(nanoparquet)
library(here)
library(scales)
library(naniar) 
library(skimr) 
library(patchwork)
library(VIM)
library(gt)
library(purrr)
library(forcats)
library(ggplot2)
library(tidyr)
library(dplyr)
```

# 1. Data Import

```{r}
# Final dataset import
wfc <- read_parquet(here("Data", "wfc_consolidated_landcover.parquet"))

# Variable changes
wfc <- wfc %>%
  mutate(
    causa = as.factor(causa),
    provincia = as.factor(provincia),
    land_cover_id = as.factor(land_cover_id),
    month_detected = factor(month_detected, levels = c("gen", "feb", "mar", "abr", "mai", "jun", 
                                                       "jul", "ago", "set", "oct", "nov", "des"))
  )

glimpse(wfc)
```

# 2. Data cleaning
## 2.1 Missing Data Analysis

```{r}
### 2.1. Overall Missingness
# Missings totals al dataframe
res_missing <- miss_var_summary(wfc)
print(res_missing)

### 2.2. Visualizing Missing Patterns
gg_miss_var(wfc) + 
  labs(title = "Missing Values by Variable")

vis_miss(wfc, warn_large_data = FALSE) +
  theme(axis.text.x = element_text(angle = 90))
```


- **Meteorology (48.5%)**: This is consistent with the historical nature of the data. Nearly half of the dataset lacks temperature records, likely due to the limited number of active weather stations during the earlier years of the study period.

- **Months (22.5%)**: This is a coding error. There are 3,246 wildfire records where the month is labeled as NA. This is likely because certain months (such as January) did not match the predefined factor levels during the transformation process.

- **Precipitation and altitude (14.2%)**: These 2,041 missing values are likely due to the technical specifications of certain Meteocat stations, which may only record temperature variables and not precipitation, or vice versa.


### 2.1.1. Month_detected

```{r}
wfc <- wfc %>%
  mutate(month_detected = month(date_detected, label = TRUE, abbr = TRUE))

wfc <- wfc %>%
  mutate(
    month_detected = month(date_detected, 
                           label = TRUE, 
                           abbr = TRUE, 
                           locale = "en_US.UTF-8")
  )

levels(wfc$month_detected)
sum(is.na(wfc$month_detected))

```

### 2.1.2 TX, TN and Termal amplitude

```{r}
# Missing patterns by county

missing_by_county <- wfc %>%
  group_by(county_clean) %>%
  summarise(
    total_fires = n(),
    n_miss_tx = sum(is.na(tx)),
    pct_miss_tx = (n_miss_tx / total_fires) * 100
  ) %>%
  arrange(desc(pct_miss_tx))

missing_by_county
```


- **Stage 1: County-level and Monthly Mean Imputation**

This step addresses records within counties that possess active weather stations but lack data for specific dates due to isolated station failures or temporary technical issues.

```{r}
# 2.1.4. Hierarchical Imputation - Stage 1: County & Month Mean
wfc_imputed <- wfc %>%
  group_by(county_clean, month_detected) %>%
  mutate(
    # Omplim TX si és NA amb la mitjana del mateix mes i comarca
    tx = ifelse(is.na(tx), mean(tx, na.rm = TRUE), tx),
    tn = ifelse(is.na(tn), mean(tn, na.rm = TRUE), tn),
    thermal_amplitude = ifelse(is.na(thermal_amplitude), mean(thermal_amplitude, na.rm = TRUE), thermal_amplitude)
  ) %>%
  ungroup()

sum(is.na(wfc_imputed$tx))
```
```{r}
# Missing patterns by County

missing_by_county <- wfc_imputed %>%
  group_by(county_clean) %>%
  summarise(
    total_fires = n(),
    # TX
    n_miss_tx = sum(is.na(tx)),
    pct_miss_tx = (n_miss_tx / total_fires) * 100,
    # TN
    n_miss_tn = sum(is.na(tn)),
    pct_miss_tn = (n_miss_tn / total_fires) * 100,
    # Thermal Amplitude
    n_miss_ta = sum(is.na(thermal_amplitude)),
    pct_miss_ta = (n_miss_ta / total_fires) * 100
  ) %>%
  arrange(desc(pct_miss_tx))

print(missing_by_county)
```

- **Stage 2: Spatio-Temporal K-Nearest Neighbors (KNN) Imputation**

For the 22 'blind' counties (those with a 100% missingness rate), we will apply the K-Nearest Neighbors (KNN) algorithm. This method identifies the 'k' most similar fire events (e.g., k=5) based on their geographic proximity latitude and longitude and assigns the average value of these neighbors to the missing record. 

```{r}
# 1.Prepare the data for imputation
wfc_prep <- wfc_imputed %>%
  mutate(
    # Convertim la data a un número (dies) per poder calcular distàncies temporals
    date_numeric = as.numeric(date_detected)
  )

# 2. Apply KNN considering Latitude, Longitude, and Date
# kNN automatically scales variables to calculate Gower's distance
wfc_final <- kNN(wfc_prep, 
                 variable = c("tx", "tn", "thermal_amplitude"), 
                 dist_var = c("latitude", "longitude", "date_numeric"), 
                 k = 5, 
                 imp_var = FALSE)

# 3. Clean up the auxiliary column
wfc_final <- wfc_final %>% select(-date_numeric)

# Final check for remaining missing values
sum(is.na(wfc_final$tx))
```

### 2.1.3 Altitude and ppt

```{r}
# Missing patterns by County

missing_by_county2 <- wfc_final %>%
  group_by(county_clean) %>%
  summarise(
    total_fires = n(),
    # PPT: 
    n_miss_ppt = sum(is.na(ppt)),
    pct_miss_ppt = (n_miss_ppt / total_fires) * 100,
    # Altitude
    n_miss_alt = sum(is.na(altitude_z)),
    pct_miss_alt = (n_miss_alt / total_fires) * 100
  ) %>% 
  arrange(desc(pct_miss_alt))

print(missing_by_county2)
```

- **Stage 1: County-level and Monthly Mean Imputation**

This step addresses records within counties that possess active weather stations but lack data for specific dates due to isolated station failures or temporary technical issues.

```{r}
wfc_imputed2 <- wfc_final %>%
  group_by(county_clean, month_detected) %>%
  mutate(
    ppt = ifelse(is.na(ppt), mean(ppt, na.rm = TRUE), ppt),
    altitude_z = ifelse(is.na(altitude_z), mean(altitude_z, na.rm = TRUE), altitude_z),
  ) %>%
  ungroup()

sum(is.na(wfc_imputed2$ppt))
```

```{r}
# Missing patterns by County
missing_by_county2 <- wfc_imputed2 %>%
  group_by(county_clean) %>%
  summarise(
    total_fires = n(),
    # PPT
    n_miss_ppt = sum(is.na(ppt)),
    pct_miss_ppt = (n_miss_ppt / total_fires) * 100,
    # Altitude
    n_miss_alt = sum(is.na(altitude_z)),
    pct_miss_alt = (n_miss_alt / total_fires) * 100
  ) %>%
  arrange(desc(pct_miss_alt))

print(missing_by_county2)
```

- **Stage 2: Spatio-Temporal KNN Imputation (to clean the 100% missing counties)**

```{r}
wfc_prep <- wfc_imputed2 %>%
  mutate(date_numeric = as.numeric(date_detected))

wfc_final2 <- kNN(wfc_prep, 
                 variable = c("ppt", "altitude_z"), 
                 dist_var = c("latitude", "longitude", "date_numeric"), 
                 k = 5, 
                 imp_var = FALSE)

# 3. Final cleanup and verification
wfc_final2 <- wfc_final2 %>% select(-date_numeric)

sum(is.na(wfc_final2$ppt))
sum(is.na(wfc_final2$altitude_z))
```

```{r}
# Missings totals al dataframe
res_missing <- miss_var_summary(wfc_final2)
print(res_missing)
```

By implementing the two-stage hierarchical imputation protocol, the overall missing data rate across the entire dataframe was successfully reduced from 6.8% to 0.55%. This substantial improvement ensures the structural integrity of the dataset for the modeling phase. The remaining 0.55% of missing values are exclusively located in the land_cover_id variable, which was not included in the spatial-temporal KNN process to avoid introducing categorical bias.


## 2.2 Preliminary variable filtering

```{r}
wfc_final2$year <- as.factor(wfc_final2$year)
wfc_final2$motivacion <- as.factor(wfc_final2$motivacion)
str(wfc_final2)
summary(wfc_final2)
```
At this stage, all remaining variables were retained for subsequent analyses. No further feature elimination was performed, as these variables were considered potentially relevant for the exploratory and modeling phases.


# 3. Exploratory Data Analysis (EDA)
## 3.1 Descriptive analysis

```{r}
### 3.1. Detailed Summary Table
skim(wfc_final2)
```

```{r}
num_summary <- wfc_final2 %>%
  select(where(is.numeric)) %>%
  map_dfr(function(x) {
    tibble(
      n = sum(!is.na(x)),
      missing = sum(is.na(x)),
      mean = mean(x, na.rm = TRUE),
      median = median(x, na.rm = TRUE),
      min = min(x, na.rm = TRUE),
      max = max(x, na.rm = TRUE)
    )
  }, .id = "variable") %>%
  mutate(
    range = paste0(round(min, 2), " - ", round(max, 2))
  ) %>%
  select(variable, n, missing, mean, median, range)

gt_num <- num_summary %>%
  mutate(variable = paste0("**", variable, "**")) %>%
  gt() %>%
  tab_header(
    title = "Descriptive statistics - Numerical variables"
  ) %>%
  fmt_markdown(columns = variable) %>%
  fmt_number(
    columns = c(mean, median),
    decimals = 2
  ) %>%
  opt_table_font(
    font = list(
      gt::google_font("Times New Roman"),
      "Times New Roman",
      "serif"
    )
  )
gt_num
```


```{r}
# 1. Crear el resum per a variables categòriques
cat_summary <- wfc_final2 %>%
  select(where(~is.character(.) | is.factor(.))) %>%
  names() %>%
  map_df(function(var) {
    wfc_final2 %>%
      count(!!sym(var)) %>%
      mutate(
        variable = var,
        level = as.character(!!sym(var)),
        percent = n / sum(n) * 100
      ) %>%
      mutate(level = ifelse(is.na(level), "Missing", level)) %>%
      select(variable, level, n, percent)
  })

# 2. Crear la taula gt

# 1. Preparem les dades fora de la taula per evitar errors de format
cat_summary_clean <- cat_summary %>%
  mutate(
    # Assegurem que el percentatge sigui numèric per evitar errors de format
    percent = as.numeric(ifelse(is.na(percent), 0, percent)),
    level = ifelse(is.na(level), "Missing", level)
  )

# 2. Generem la taula
gt_cat <- cat_summary_clean %>%
  gt(groupname_col = "variable") %>%
  tab_header(
    title = "Descriptive statistics - Categorical variables"
  ) %>%
  # Forçar negreta als noms de les variables (grups)
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  # Forçar negreta a les capçaleres de les columnes
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  # CONFIGURACIÓ DE DECIMALS: 2 decimals per a la columna percent
  fmt_number(
    columns = percent, 
    decimals = 2
  ) %>%
  # Estil tipogràfic Times New Roman
  opt_table_font(
    font = list(
      gt::google_font("Times New Roman"),
      "Times New Roman",
      "serif"
    )
  )

# Visualització
gt_cat
```


# 4. Univariate Analysis (Distributions)
## 4.1. Numeric Variables (Outliers check)

This code uses pivot_longer to prepare all numerical data at once and facet_wrap to create a grid of plots, which is much more efficient than doing them one by one.

```{r packages import, warning=FALSE, message=FALSE}
# Prepare numerical data in long format
wfc_num_long <- wfc_final2 %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

# 1.1. Combined Histogram and Density Plots
# Useful for checking the distribution shape (normality, skewness)
ggplot(wfc_num_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "steelblue", alpha = 0.6) +
  geom_density(color = "firebrick", size = 1) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(
    title = "Distribution Analysis: Histograms and Density",
    subtitle = "Numerical variables from the wildfire dataset",
    x = "Value", 
    y = "Density"
  )

# 1.2. Boxplots
# Crucial for identifying outliers and understanding the interquartile range (IQR)
ggplot(wfc_num_long, aes(x = variable, y = value, fill = variable)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.alpha = 0.5) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Outlier Detection: Boxplots",
    x = "Variable", 
    y = "Value"
  )
```

## 4.2. Categorical Variables (Frequency)

```{r}
# Create a list of bar plots for all categorical variables
# Using Lapply to iterate through character and factor columns
categorical_vars <- wfc_final2 %>%
  select(where(~is.character(.) | is.factor(.))) %>%
  names()

plots_cat <- lapply(categorical_vars, function(var) {
  ggplot(wfc_final2, aes(y = fct_infreq(!!sym(var)))) +
    geom_bar(fill = "darkseagreen", alpha = 0.8) +
    theme_minimal() +
    labs(
      title = paste("Frequency Distribution:", var),
      x = "Count (Number of Fires)", 
      y = "Category"
    )
})

walk(plots_cat, print)


```

