---
title: "ETL_proces"
author: "Gemma Bargalló Solé"
date: "2026-03-31"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      position: left
    toc_depth: 5   
  word_document: default
  pdf_document: default
---
```{r packages import, warning=FALSE, message=FALSE}
library(DataExplorer)
library(dplyr)
library(ggplot2)
library(gt)
library(stringr) 
library(forcats)
library(lubridate)
library(sf)
library(tidyverse)
library(janitor)
library(here)
library(readxl)
library(httr)
library(fs)
library(terra)
library(arrow)
```

# 1. CATALONIA WILDFIRES DATASET
## 1.1 DATA IMPORT

We start by working with the data on Catalonia Wildfires. We observe a summary to see the size and structure of our data.

```{r data import}
ruta_fitxer <- here("Data", "WildfiresCat.xlsx")
wfc<- read_xlsx(ruta_fitxer, sheet = 1, guess_max = 30000)
summary(wfc)
str(wfc)
```
The dataset has 31,319 rows (or observations) and 31 columns (or variables).

After importing the dataset, a preliminary inspection using summary() and str() reveals several structural inconsistencies that must be addressed before any analytical or modeling tasks can begin.

As observed in the output above, there is a significant discrepancy between the conceptual nature of the variables and their storage mode in R:
- **Type mismatch**: Most variables, including those representing physical quantities (e.g., `SuperficieArbolada`, `CoordenadaX`) and temporal markers, have been imported as character strings instead of numeric or Date objects.
- **Localization issues**: Numeric values in the Catalan dataset use the comma (,) as a decimal separator, which prevents R from recognizing them as double-precision numbers.
- **Categorical data**: Variables such as Causa, Provincia, and AfectoEspacioProtegido are currently treated as plain text rather than factors, which would be more efficient for statistical grouping and Machine Learning.
- **Temporal formatting**: The columns Detectado and Extinguido contain date and time information in a non-standard string format that requires parsing for time-series analysis.

## 1.2 DATA TRANFORMATION AND CLEANING (ETL)

To resolve these issues, we will implement a transformation pipeline using the tidyverse ecosystem. This stage aims to standardize the dataset and ensure that each column reflects its true statistical nature.

```{r warning=FALSE, message=FALSE}
# 1. Clean column names to snake_case for easier coding
wfc <- wfc %>% 
  clean_names()

# 2. Convert character strings to numeric by fixing decimal separators
# We select the columns that represent areas and coordinates
wfc <- wfc %>%
  mutate(across(
    c(superficie_arbolada, superficie_no_arbolada, superficie_total_forestal, 
      superficie_agricola, otras_superficies_noforestales, 
      coordenada_x, coordenada_y),
    ~ as.numeric(str_replace(., ",", "."))
  ))

# 3. Parse and split Date/Time columns
wfc <- wfc %>%
  mutate(
    detectado = dmy_hms(detectado),
    extinguido = dmy_hms(extinguido),
    year = year(detectado) # Extract year for temporal analysis
  )

# 4. Convert key binary and categorical variables into factors
wfc <- wfc %>%
  mutate(across(
    c(provincia, municipio,  comarca_isla, causa, afecto_espacio_protegido, afecto_tierras_agrarias, afecto_zar, afecto_zonas_interfaz_urbano_forestal, numero_municipios_afectados),
    as.factor
  ))

# Review the cleaned structure
glimpse(wfc)

```


### 1.2.1 Missing values analysis and dimensionality reduction 

Before proceeding with geospatial transformations, it is essential to evaluate the completeness of our dataset. High-density NA columns or administrative variables that do not contribute to the predictive modeling of wildfires should be removed to optimize the pipeline.

```{r warning=FALSE, message=FALSE}
# 1. Visualizing the percentage of missing values per variable
plot_missing(wfc)

# 2. Calculating the exact percentage of NAs
missing_summary <- wfc %>%
  summarise(across(everything(), ~ sum(is.na(.)) / n() * 100)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") %>%
  arrange(desc(pct_missing))

print(missing_summary)
```

Based on the diagnostic above and the objectives of this TFM, we will drop columns that fall into the following categories:

- **Empty columns**: Variables with near 100% missing values (`entidad_menor`, `datum` and `numero_parte_pss`).
- **Formatting/administrative junk**: Columns containing only "invisible" characters or internal migration codes that lack physical or environmental meaning (e.g., hoja, cuadricula, huso, estado).
- **Redundant labels**: Broad administrative levels like `comunidad` (all are "CATALUÑA").

```{r}
# We define the list of variables to keep or drop
wfc_reduced <- wfc %>%
  select(
    # Keepers: Temporal & Administrative
    campania, provincia, comarca_isla, municipio,
    # Keepers: Coordinates & Dates
    coordenada_x, coordenada_y, detectado, extinguido, year,
    # Keepers: Fire Dynamics (Target variables for ML)
    causa, motivacion, starts_with("superficie_"), otras_superficies_noforestales, 
    # Keepers: Environmental/Social impacts
    afecto_zonas_interfaz_urbano_forestal, afecto_espacio_protegido, 
    afecto_tierras_agrarias, afecto_zar
  ) 

# Final check of the dimensions
dim(wfc_reduced)
```


### 1.2.2 Handling missing values and geospatial preparation

A critical observation from the previous step is the presence of NA values in the coordinate columns, particularly in older records (e.g., the 1968 campaign). Since our objective involves spatial modeling and mapping, records without valid geographic coordinates cannot be used for spatial joins with meteorological data.

Additionally, we notice that some columns contain empty strings represented by a special character (invisible spaces or " "), which should be treated as true NA values to maintain data integrity.


```{r}
# 1. Convert suspicious empty strings or " " into real NAs
# 2. Filter out records without coordinates 
wfc_spatial <- wfc_reduced %>%
  mutate(across(where(is.character), ~na_if(str_trim(.), ""))) %>%
  filter(!is.na(coordenada_x) & !is.na(coordenada_y))

# 3. Quick check: how many records did we keep?
n_original <- nrow(wfc_reduced)
n_spatial <- nrow(wfc_spatial)
lost_records <- n_original - n_spatial

print(paste("Original records:", n_original))
print(paste("Spatial records remaining:", n_spatial))
print(paste("Records lost due to missing coordinates:", lost_records))
```

### 1.2.3 Coordinate Reference System (CRS) transformation

The coordinates in the Catalan dataset are originally projected in ETRS89 / UTM zone 31N (EPSG:25831). However, to ensure compatibility with global datasets (like our Spanish wildfire data or meteorological layers) and for interactive visualization in Leaflet, we need to transform them into geographic coordinates WGS84 (EPSG:4326).

```{r}
# 1. Convert the data frame into a spatial object (sf)
# We use the original UTM 31N projection (EPSG:25831)
wfc_sf <- st_as_sf(wfc_spatial, 
                   coords = c("coordenada_x", "coordenada_y"), 
                   crs = 25831)

# 2. Transform the projection to WGS84 (Lat/Lon)
wfc_sf_wgs84 <- st_transform(wfc_sf, crs = 4326)

# 3. Extract the new Lat/Lon back into columns for standard data frame use
wfc_final <- wfc_sf_wgs84 %>%
  mutate(longitude = st_coordinates(.)[,1],
         latitude = st_coordinates(.)[,2]) %>%
  st_drop_geometry() # Keep it as a data frame for easier merging later

# Final inspection of the new coordinates
wfc_final %>% select(campania, municipio, longitude, latitude) %>% head()
```

Let's see again the final missing values of our dataset:

```{r}
colSums(is.na(wfc_final))
```

#### Refinement of temporal variables

```{r}
wfc_final <- wfc_final %>%
  mutate(
    detectado = as.POSIXct(detectado),
    extinguido = as.POSIXct(extinguido)
  )

# 2. Extract separate Date and Time columns
wfc_final <- wfc_final %>%
  mutate(
    # Extraction of the Date component
    date_detected = as.Date(detectado),
    date_extinguished = as.Date(extinguido),
    
    # Extraction of the Time component (HH:MM:SS)
    time_detected = format(detectado, "%H:%M:%S"),
    time_extinguished = format(extinguido, "%H:%M:%S"),
    
    # Extract numerical Hour for later statistical visualization (0-23)
    hour_detected = hour(detectado),
    month_detected = month(detectado, label = TRUE)
  )
  
    
# 3. Verify the new structure
wfc_final %>% 
  select(date_detected, time_detected, date_extinguished, time_extinguished) %>% 
  head()

wfc_final <- wfc_final %>%
  select(-detectado, -extinguido)
```



In summary, the current dataset has been consolidated and reviewed, confirming that the temporal series analyzed covers the period from **1998 to 2022**.



# 2. SPAIN WILDFIRES DATASET
## 2.1 DATA IMPORT

Let's continue working with the data on Spain Wildfires. We observe a summary to see the size and structure of our data.

```{r data import, warning=FALSE, message=FALSE}
ruta_fitxer <- here("Data", "WildfiresSpain.csv")
wfs<- read_csv(ruta_fitxer, guess_max = 30000)
summary(wfs)
str(wfs)
```
The dataset has 292,181 rows (or observations) and 21 columns (or variables).

The inspection of the national-level wildfire dataset (1968–2023) reveals a more complex landscape regarding data integrity. While the technical importation via read_csv() correctly identified most data types (numeric and Date objects), several critical issues emerge from the structural analysis:

- **Significant missingness (NA's)**: Unlike the regional Catalan dataset, there is a substantial volume of missing values in key analytical fields.
- **Temporal scope and filtering**: The dataset spans from 1968 to 2023. To maintain consistency with the established research period (1998–2022), a significant subset of the data must be truncated, ensuring that long-term trends are not skewed by the lower reporting standards of the pre-1990s era.
- **Logical inconsistencies**: The summary reveals anomalies in operational timeframes, such as negative values in time_ctrl and time_ext (minimums of -1430 and -1352 respectively). These represent physical impossibilities that suggest data entry errors or corruption, requiring a robust outlier removal protocol.
- **Categorical encoding**: Although administrative units (idprovincia, idcomunidad) and causa are stored as numeric doubles, they represent discrete categories. These must be converted into factors to allow the Random Forest algorithm to correctly interpret them as qualitative features rather than continuous variables.

## 2.2 DATA TRANFORMATION AND CLEANING (ETL)

To resolve these issues, we will implement a transformation pipeline using the tidyverse ecosystem. This stage aims to standardize the dataset and ensure that each column reflects its true statistical nature.

```{r warning=FALSE, message=FALSE}
# 1. Clean column names to snake_case for consistency
wfs <- wfs %>% 
  clean_names()

# 2. Parse dates and extract temporal components
wfs <- wfs %>%
  mutate(
    fecha = as.Date(fecha),
    year = year(fecha),
    month = month(fecha)
  )

# 3. Standardize numeric variables
# Ensuring surface and coordinates are numeric (handling potential string issues)
wfs <- wfs %>%
  mutate(across(
    c(superficie, lat, lng),
    ~ as.numeric(as.character(.))
  ))

# 4. Handle logical inconsistencies in time columns
# Converting negative durations (errors) to NA to avoid biasing the analysis
wfs <- wfs %>%
  mutate(
    time_ctrl = ifelse(time_ctrl < 0, NA, time_ctrl),
    time_ext = ifelse(time_ext < 0, NA, time_ext)
  )

# 5. Convert categorical identifiers and binary flags into factors
wfs <- wfs %>%
  mutate(across(
    c(idcomunidad, idprovincia, idmunicipio, municipio, 
      causa, causa_supuesta, causa_desc, latlng_explicit),
    as.factor
  ))

# Review the cleaned structure
glimpse(wfs)

```

### 2.2.1 GEOGRAPHIC SUBSETTING: COMMUNITY SELECTION

To narrow the scope of our analysis to a specific region, we filter the dataset to include only observations from the community identified with the ID 2, Catalonia. This allows for a more localized and detailed study of fire patterns.

```{r}
# Filter the dataset by 'idcomunidad'
wfs <- wfs %>%
  filter(idcomunidad == "2")

# Optional: Drop unused factor levels to keep the data clean
wfs$idcomunidad <- droplevels(wfs$idcomunidad)

# Verify the number of observations in the subset
nrow(wfs)
```
The resulting subset consists of 10,161 entries, representing all recorded incidents for Community ID 2.


### 2.2.2 Missing values analysis and dimensionality reduction 

Before proceeding with geospatial transformations, it is essential to evaluate the completeness of our dataset. High-density NA columns or administrative variables that do not contribute to the predictive modeling of wildfires should be removed to optimize the pipeline.

```{r warning=FALSE, message=FALSE}
# 1. Visualizing the percentage of missing values per variable
plot_missing(wfs)

# 2. Calculating the exact percentage of NAs
missing_summary <- wfs %>%
  summarise(across(everything(), ~ sum(is.na(.)) / n() * 100)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") %>%
  arrange(desc(pct_missing))

print(missing_summary)
```

Based on the diagnostic above and the objectives of this TFM, we will drop columns that fall into the following categories:

- **Empty columns**: Variables with near 100% missing values 
- **Redundant labels**: Broad administrative levels like `idcomunidad`.

```{r warning=FALSE, message=FALSE}
# We define the list of variables to keep or drop
wfs <- wfs %>%
  select(-idcomunidad, -latlng_explicit)

# Final check of the dimensions
dim(wfs)
```

### 2.2.3 Handling missing values and geospatial preparation

A critical observation from the previous step is the presence of NA values in the coordinate columns, particularly in older records. Since our objective involves spatial modeling and mapping, records without valid geographic coordinates cannot be used for spatial joins with meteorological data.


```{r}
# 1. Filter out records without coordinates 
wfs_spatial <- wfs %>%
  filter(!is.na(lat) & !is.na(lng))

# 2. Quick check: how many records did we keep?
n_original <- nrow(wfs)
n_spatial <- nrow(wfs_spatial)
lost_records <- n_original - n_spatial

print(paste("Original records:", n_original))
print(paste("Spatial records remaining:", n_spatial))
print(paste("Records lost due to missing coordinates:", lost_records))
```
Let's see again the final missing values of our dataset:

```{r}
colSums(is.na(wfs_spatial))
```

In summary, the current dataset has been consolidated and reviewed, confirming that the temporal series analyzed covers the period from **1983 to 2020**.


# 3. DATASET COMPATIBILITY AND SELECTION STRATEGY

Before proceeding with the exploratory analysis, a comparative assessment of the available data sources was conducted. We currently have two potential datasets for Catalonia:

- **Regional dataset (Catalonia)**: 14,397 observations, 24 variables (1998–2022).

- **National dataset (Spain - filtered)**: 5,404 observations, 21 variables (1983–2020).

After evaluation, we have decided not to merge the datasets for the following reasons:

- **Risk of redundancy**: There is a high probability that the National dataset is a partial subset of the Regional one for the overlapping years (1998–2020), which would lead to double-counting incidents.

- **Temporal and structural inconsistency**: The datasets differ significantly in volume and time range. Merging them without a verified unique identifier (UID) would compromise the statistical reliability of the trends.

- **Data richness**: The Regional Dataset (Catalonia) is selected as the primary source because it offers a higher density of observations, more specific local variables, and more recent data (up to 2022).

**Conclusion**: To ensure maximum precision and avoid bias, the analysis will focus exclusively on the Regional Dataset (Catalonia), using the National data only as an auxiliary reference if specific historical validation (pre-1998) is required.


# 4. METEOROLOGICAL DATASETS
## 4.1 DATA IMPORT

```{r warning=FALSE, message=FALSE}
data_folder <- "Data"
raw_weather_files <- dir_ls(data_folder, glob = "*.txt")

process_weather_file <- function(file_path) {
  all_lines <- read_lines(file_path)
  
  # --- METADATA ---
  county_line <- all_lines[str_detect(all_lines, "Comarca:")]
  county_val  <- str_trim(str_remove(county_line, "Comarca:"))
  
  variable_lines <- all_lines[str_detect(all_lines, "Variable")]
  all_vars       <- str_remove(variable_lines, "Variable[0-9]*:") %>% 
                   str_trim() %>% 
                   str_flatten(collapse = " | ")
  
  extract_coord <- function(line) {
    val <- str_extract(line, "(?<=: ).*(?= m)")
    return(as.numeric(str_trim(val)))
  }
  
  x_coord <- extract_coord(all_lines[str_detect(all_lines, "X UTM31")])
  y_coord <- extract_coord(all_lines[str_detect(all_lines, "Y UTM31")])
  z_alt   <- extract_coord(all_lines[str_detect(all_lines, "Z UTM31")])
  
  # --- DATA ---
  header_index <- which(str_detect(all_lines, "^ANY"))
  if (length(header_index) == 0) return(NULL)
  
  # FORCEM que totes les columnes siguin CHARACTER per evitar el conflicte Double vs Character
  df <- read_table(file_path, skip = header_index - 1, 
                   show_col_types = FALSE, 
                   col_types = cols(.default = "c"))
  
  df <- df %>%
    mutate(
      county      = county_val[1],
      variables   = all_vars,
      utm_x       = x_coord[1],
      utm_y       = y_coord[1],
      altitude_z  = z_alt[1],
      source_file = path_file(file_path)
    )
  
  return(df)
}

# Unim tots els fitxers 
weather_master_df <- raw_weather_files %>%
  map(process_weather_file) %>% 
  bind_rows()

# --- NETEJA POST-UNIÓ ---
weather_master_df <- weather_master_df %>%
  # Convertim a número només el que realment ho és, gestionant errors automàticament
  mutate(across(c(ANY, MES, DIA, any_of(c("PPT", "TX", "TN", "INS", "TM"))), 
                ~as.numeric(str_replace(., ",", ".")))) %>% 
  mutate(date_detected = as.Date(paste(ANY, MES, DIA, sep = "-"))) %>%
  select(date_detected, county, everything())

write_parquet(weather_master_df, "Data/consolidated_weather_catalonia.parquet")

```

```{r warning=FALSE, message=FALSE}
ruta_fitxer <- here("Data", "consolidated_weather_catalonia.parquet")
wc<- read_parquet(ruta_fitxer)
summary(wc)
str(wc)
```
The meteorological component of this study is derived from 72 raw weather stations' records across Catalonia, covering an extensive historical period from 1950 to 2024. After consolidating these heterogeneous text files into a single unified structure, the resulting dataset consists of 1,972,337 daily observations and 14 variables.

A preliminary inspection using summary() and glimpse() reveals the initial state of the data:

- **Massive temporal scale**: The data spans over 70 years, providing a robust baseline for long-term wildfire climate analysis. However, there is a high frequency of missing values (NAs) in specific indicators like Maximum Temperature (TX) and Minimum Temperature (TN).

- **Variable fragmentation**: The dataset contains diverse physical quantities, including daily accumulated precipitation (PPT), maximum/minimum temperatures (TX/TN), and solar insolation (INS). These variables were originally scattered across different files and have been successfully merged using a relational key based on date and county.

- **Geospatial metadata**: Each observation is now enriched with precise geographic attributes extracted from the original file headers, including UTM coordinates (X, Y) and altitude (Z). 

- **Data consistency challenges: *Type standardization**: While the initial import treated many columns as strings to avoid conflicts, they have been successfully cast into numeric and Date objects to allow for mathematical computation.
  - Inconsistent indicators: Variables like INS (Insolation) show extreme sparsity (over 1.7 million NAs), suggesting they may only be available for specific stations or recent time periods.
  - Outlier detection: Initial summaries show some extreme values (e.g., -99.9 in PPT or INS), which likely act as internal placeholders for "null" or "error" in the original station sensors and will require further filtering.
  
  
## 4.2 DATA TRANSFORMATION AND CLEANING (ETL)

To resolve these issues, we will implement a transformation pipeline using the tidyverse ecosystem. This stage aims to standardize the dataset and ensure that each column reflects its true statistical nature.

```{r}
# 1. Clean column names to snake_case for consistency
weather_cat <- weather_master_df %>% 
  clean_names()

# 2. Convert character strings to numeric and fix placeholders
# Note: Meteocat files often use codes like -99.9 or -999.9 for missing data.
# We convert these to actual NAs during the process.
weather_cat <- weather_cat %>%
  mutate(across(
    c(ppt, tx, tn, ins, utm_x, utm_y, altitude_z),
    ~ as.numeric(.)
  )) %>%
  mutate(across(
    c(ppt, tx, tn, ins),
    ~ ifelse(. < -90, NA, .) 
  ))

# 3. Consolidate Date and Time analysis
# We already have date_detected as a Date object, but we extract components 
# to ensure compatibility with the wildfire dataset
weather_cat <- weather_cat %>%
  mutate(
    year = year(date_detected),
    month = month(date_detected, label = TRUE, abbr = FALSE),
    day_of_week = wday(date_detected, label = TRUE)
  )

# 4. Convert categorical variables into factors
# This is key for grouping and later for the Random Forest model
weather_cat <- weather_cat %>%
  mutate(across(
    c(county, variables, source_file),
    as.factor
  ))

# 5. Feature Engineering: Calculate Thermal Amplitude
# A high difference between Max and Min temp often correlates with fire risk
weather_cat <- weather_cat %>%
  mutate(thermal_amplitude = tx - tn)

# Review the cleaned structure of the weather dataset
glimpse(weather_cat)
```


### 4.2.1 Missing values analysis and dimensionality reduction 

Before proceeding with geospatial transformations, it is essential to evaluate the completeness of our dataset. High-density NA columns or administrative variables that do not contribute to the predictive modeling of wildfires should be removed to optimize the pipeline.

```{r warning=FALSE, message=FALSE}
# 1. Visualizing the percentage of missing values per variable
plot_missing(weather_cat)

# 2. Calculating the exact percentage of NAs
missing_summary <- weather_cat %>%
  summarise(across(everything(), ~ sum(is.na(.)) / n() * 100)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") %>%
  arrange(desc(pct_missing))

print(missing_summary)
```


Based on the diagnostic above and the objectives of this TFM, we will drop columns that fall into the following categories:

- **Empty columns**: Variables with near 100% missing values 
- **Redundant labels**: Broad administrative levels.

```{r warning=FALSE, message=FALSE}
# We define the list of variables to keep or drop
weather_cat <- weather_cat %>%
  select(-variables, -source_file, -ins)

# Final check of the dimensions
dim(weather_cat)
```
Since the wildfire dataset for Catalonia contains records from 1998 to 2020, we have filtered the meteorological data to include only the period between January 1st, 1998, and December 31st, 2020, ensuring full temporal alignment between both data sources.

```{r}
# Filtering the weather dataset to match the wildfire temporal scope
weather_cat <- weather_cat %>%
  filter(date_detected >= as.Date("1998-01-01") & 
         date_detected <= as.Date("2020-12-31"))
dim(weather_cat)
```
let's see again the NA's total.

### 4.2.2 Coordinate Reference System (CRS) transformation

```{r}
# 1. Convert the dataframe into a spatial object (sf)
# We specify the input columns (utm_x, utm_y) and the CRS (EPSG:25831 for UTM 31N)
weather_sf <- st_as_sf(weather_cat, 
                       coords = c("utm_x", "utm_y"), 
                       crs = 25831, 
                       remove = FALSE) # Keep original UTM columns for the model

# 2. Transform the Coordinate Reference System (CRS) to WGS84 (Latitude/Longitude)
# EPSG:4326 is the standard used by GPS and Google Maps
weather_sf_transformed <- st_transform(weather_sf, crs = 4326)

# 3. Extract the new coordinates into separate numeric columns
coords_lonlat <- st_coordinates(weather_sf_transformed)

weather_cat <- weather_cat %>%
  mutate(
    longitude = coords_lonlat[,1],
    latitude  = coords_lonlat[,2]
  )

# Verify the result
weather_cat %>% 
  select(county, utm_x, utm_y, longitude, latitude) %>% 
  head()
```


```{r}
# Refined mapping to ensure 100% match with the weather dataset
wfc_final <- wfc_final %>%
  mutate(county_clean = case_match(comarca_isla,
    "ALT EMPORDA"      ~ "ALT EMPORDÀ",
    "BAIX PENEDES"     ~ "BAIX PENEDÈS",
    "ALT PENEDES"      ~ "ALT PENEDÈS",
    "ALTA RIBAGORZA"   ~ "ALTA RIBAGORÇA",
    "BAIX EMPORDA"     ~ "BAIX EMPORDÀ",
    "BARCELONES"       ~ "BARCELONÈS",
    "BERGEDA"          ~ "BERGUEDÀ",
    "CONCA DE BARBERA" ~ "CONCA DE BARBERÀ",
    "EL GIRONES"       ~ "GIRONÈS",
    "EL SEGRIA"        ~ "SEGRIÀ",
    "L URGELL"         ~ "URGELL",      # Matching 'L URGELL' to 'URGELL' (Weather DS)
    "LA GARROTXA"      ~ "GARROTXA",
    "LA NOGERA"        ~ "NOGUERA",
    "LA SEGARRA"       ~ "SEGARRA",
    "LA SELVA"         ~ "SELVA",
    "LES GARRIGES"     ~ "GARRIGUES",
    "MONSIA"           ~ "MONTSIÀ",
    "PALLARS JUSSA"    ~ "PALLARS JUSSÀ",
    "PALLARS SOBIRA"   ~ "PALLARS SOBIRÀ",
    "PLA D ESTANY"     ~ "PLA DE L'ESTANY",
    "RIBERA D EBRE"    ~ "RIBERA D'EBRE",
    "RIPOLLES"         ~ "RIPOLLÈS",
    "SOLSONES"         ~ "SOLSONÈS",
    "TARRAGONES"       ~ "TARRAGONÈS",
    "VAL D ARAN"       ~ "VAL D'ARAN",
    "VALLS OCCIDENTAL" ~ "VALLÈS OCCIDENTAL",
    "VALLS ORIENTAL"   ~ "VALLÈS ORIENTAL",
    "PLA D URGELL"     ~ "URGELL",      # Consolidating both into 'URGELL'
    .default = as.character(comarca_isla)
  )) %>%
  # Standardize casing and remove whitespace to avoid "Ghost Mismatches"
  mutate(county_clean = str_trim(toupper(county_clean)))

# Also standardize the weather dataset names
weather_cat <- weather_cat %>%
  mutate(county = str_trim(toupper(county)))
```


## 4.3 JOINING DATASETS

```{r}
# 1. Verification of Key Consistency
# Ensure both datasets have the joining keys in the same format (Uppercase & Trimmed)
wfc_final <- wfc_final %>%
  mutate(county_clean = str_trim(toupper(county_clean)))

weather_cat <- weather_cat %>%
  mutate(county = str_trim(toupper(county)))

# --- OPTION A: AGGREGATE WEATHER DATA BY COUNTY AND DATE ---
# This step prevents row duplication by calculating the daily mean of all 
# stations within the same county.
weather_daily_avg <- weather_cat %>%
  group_by(county, date_detected) %>%
  summarise(
    # Core weather variables (mean of all stations in the county)
    ppt = mean(ppt, na.rm = TRUE),
    tx  = mean(tx, na.rm = TRUE),
    tn  = mean(tn, na.rm = TRUE),
    thermal_amplitude = mean(thermal_amplitude, na.rm = TRUE),
    # Spatial metadata (average altitude of stations in the county)
    altitude_z = mean(altitude_z, na.rm = TRUE),
    .groups = "drop"
  )

# 3. Executing the Left Join
# We use wfc_final as the base (left) to preserve all wildfire records.
# The join is performed on both the Spatial (County) and Temporal (Date) dimensions.
wfc_consolidated <- wfc_final %>%
  left_join(weather_daily_avg, 
            by = c("county_clean" = "county", 
                   "date_detected" = "date_detected"))

# 3. Quality Control: Identifying Missing Matches
# We check how many wildfires did not find a corresponding weather record
missing_weather_count <- sum(is.na(wfc_consolidated$ppt))
total_fires <- nrow(wfc_consolidated)

message(paste("Total Wildfire Records:", total_fires))
message(paste("Records without Weather Data:", missing_weather_count))
message(paste("Success Rate:", round((1 - (missing_weather_count / total_fires)) * 100, 2), "%"))

# 4. Cleanup: Remove redundant or technical columns if necessary
# For example, removing the 'source_file' or 'variables' columns from the weather data
wfc_consolidated <- wfc_consolidated %>%
  select(-any_of(c("any", "mes", "dia")))

# Preview the final merged dataset
glimpse(wfc_consolidated)
```
The core analytical dataset was generated by executing a left outer join between the wildfire event log (wfc_final) and the daily meteorological time-series (weather_cat).
- **Relational strategy**: The merge utilized a composite key consisting of the administrative county (Comarca) and the detection date. This logic ensures that each ignition event is attributed with the precise environmental stressors—such as daily precipitation (ppt), maximum temperature (tx), and thermal amplitude—present in that specific region on the day of the outbreak.
- **Spatial attribution**: By aligning the records via the standardized county_clean field, the model accounts for macro-climatic variations across Catalonia’s diverse geography, from the humid Pyrenees to the semi-arid Ebro valley.
- **Handling gaps**: The choice of a left_join ensures that no wildfire records are discarded during the merge. Observations with missing meteorological attributes (e.g., those in "Indeterminate" locations) are preserved as NA values, providing transparency regarding data coverage for the subsequent Machine Learning validation.


### 4.3.1 Data cleaning and post-processing

```{r}
wfc_consolidated <- wfc_consolidated %>%
  # 1. Remove redundant columns
  select(-campania, -comarca_isla) %>%
  
  # 2. Convert county_clean to factor for categorical analysis
  mutate(county_clean = as.factor(county_clean)) %>%
  
  # 3. Convert time strings to proper time objects (hms)
  # This allows for numerical calculations with time if needed
  mutate(
    time_detected = hms(time_detected),
    time_extinguished = hms(time_extinguished)
  )

# Verify the changes
glimpse(wfc_consolidated)
```

# 5.LAND COVER DATASET
## 5.1 DATA IMPORT

In this section, we integrate the Land Cover Maps of Catalonia (MCSC) from the ICGC. Since we have fire records from 1998 to 2022 but land cover maps are only available for specific years (2009, 2018, and 2019-2020), we will apply a **temporal proximity join**:

* **Fires (1998–2013):** Assigned to the 2009 Land Cover Map.
* **Fires (2014–2018):** Assigned to the 2018 Land Cover Map.
* **Fires (2019–2022):** Assigned to the 2019-2020 Land Cover Map.

To ensure the project is portable and efficient, we will "crop" the original heavy rasters from the external drive to the study area and save them in the local `Data/` folder using LZW compression.

```{r}
# 3. Define paths (External USB and Local Project)
#usb_path <- "E:/MASTER Data science/Màster en Data Science/TFM/"
#tif_files <- c("cobertes-sol-v1r0-2009.tif", 
#               "cobertes-sol-v1r0-2018.tif", 
#               "cobertes-sol-v1r0-2019-2020.tif")

# 4. Process each raster: Crop and Compress
#for (f in tif_files) {
#  message("Processing and cropping: ", f)
  
  # Load original raster from USB
#  r_source <- rast(paste0(usb_path, f))
  
  # Project bounding box to match Raster CRS
#  bbox_proj <- st_transform(study_area_bbox, crs(r_source))
  
  # Crop the raster to the study area
#  r_cropped <- crop(r_source, vect(bbox_proj))
  
  # Save locally in the Data folder with high compression
#  writeRaster(r_cropped, here("Data", f), 
#              gdal = c("COMPRESS=LZW", "PREDICTOR=2"), 
#              overwrite = TRUE)
#}
```


```{r}
# 1. Convert to spatial format (EPSG:4326 - WGS84)
incendis_sf <- st_as_sf(wfc_consolidated, coords = c("longitude", "latitude"), crs = 4326)

# 2. Create a bounding box to crop the rasters (saves space for Git)
# We add a small buffer (0.1 degrees) around our fire points
study_area <- st_bbox(incendis_sf) %>% st_as_sfc() %>% st_buffer(0.1)
```



## 5.2 Spatial Extraction and Temporal Mapping

We now iterate through each Land Cover raster. For each file, we:

1. Clip the raster to our study area to reduce file size for GitHub compatibility.
2. Reproject our fire coordinates to match the raster's CRS (usually UTM 31N).
3. Extract the land cover class for each fire location.
4. Filter and assign the values to the main dataset based on the fire's year.

```{r}
# Initialize columns for land cover
wfc_consolidated$land_cover_id <- NA

for (f in tif_files) {
  message("Processing raster: ", f)
  
  # Load raster from USB
  r_full <- rast(paste0(usb_path, f))
  
  # Project study area to Raster CRS for cropping
  study_area_proj <- st_transform(study_area, crs(r_full))
  
  # Crop and Save locally (This allows the tutor to run it without the USB)
  r_cropped <- crop(r_full, vect(study_area_proj))
  writeRaster(r_cropped, here("Data", f), gdal="COMPRESS=LZW", overwrite=TRUE)
  
  # Project fire points to Raster CRS
  incendis_proj <- st_transform(incendis_sf, crs(r_cropped))
  
  # Extract values
  extracted_values <- extract(r_cropped, vect(incendis_proj))[, 2]
  
  # Temporal assignment logic
  if (grepl("2009", f)) {
    # Fires from start until 2013
    mask <- wfc_consolidated$year <= 2013
    wfc_consolidated$land_cover_id[mask] <- extracted_values[mask]
    
  } else if (grepl("2018", f)) {
    # Fires from 2014 to 2018
    mask <- wfc_consolidated$year >= 2014 & wfc_consolidated$year <= 2018
    wfc_consolidated$land_cover_id[mask] <- extracted_values[mask]
    
  } else if (grepl("2019-2020", f)) {
    # Fires from 2019 to 2022
    mask <- wfc_consolidated$year >= 2019
    wfc_consolidated$land_cover_id[mask] <- extracted_values[mask]
  }
}

# Save the final light-weight dataset
nanoparquet::write_parquet(wfc_consolidated, here("Data", "wfc_final_analysis.parquet"))

# Final check
table(wfc_consolidated$land_cover_id, useNA = "always")
```


MODEL D'ELEVACIONS

