---
title: "Random Forest"
author: "Gemma Bargalló Solé"
date: "2026-04-28"
html_document:
    code_folding: hide
    toc: true
    toc_float:
      position: left
    toc_depth: 5   
  word_document: default
  pdf_document: default
---


# 1. Data Import

```{r}
# Final dataset import
wfc_final2 <- read_csv(here("Data", "wfc_final2.csv"))

```

# 2. Pre-processing and Feature Engineering
## 2.1 Data Cleaning and Feature Selection
In this step, we filter the columns, handle categorical variables as factors, and prepare the dataset for the model.

```{r}
# Load required libraries
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)

# 1. Feature Selection and Formatting
# FINAL MODEL DATASET SELECTION
wfc_model_final <- wfc_final2 %>%
  select(
    # Target (Change this based on your goal)
    superficie_total_forestal, 
    
    # Location
    county_clean, altitude_z, latitude, longitude,
    
    # Meteorology
    tx, ppt, thermal_amplitude,
    
    # Time
    month_detected, hour_detected,year,
    
    # Land Cover & Risk
    land_cover_id, afecto_zonas_interfaz_urbano_forestal, 
    afecto_espacio_protegido, afecto_zar,
    
    # Context
    causa
  ) %>%
  # Convert to factors for Random Forest
  mutate(across(where(is.character), as.factor),
         land_cover_id = as.factor(land_cover_id),
         month_detected = as.factor(month_detected),
         # If hour_detected is numeric, we might keep it as is or factorize it
         hour_detected = as.numeric(hour_detected)) %>%
  drop_na()
```



## 2.2 Correlation Analysis

We use Spearman correlation because your previous histograms showed that the numerical variables are not normally distributed.

```{r}
# Select only numerical columns for the matrix
numeric_vars <- wfc_model_final %>% select(where(is.numeric))
cor_matrix <- cor(numeric_vars, method = "spearman")

# Plotting the correlation matrix
corrplot(cor_matrix, 
         method = "color", 
         type = "upper", 
         addCoef.col = "black", 
         tl.col = "black", 
         diag = FALSE,
         title = "\nSpearman Correlation Matrix of Predictors",
         mar = c(0,0,2,0))
cor_matrix
```


## 2.3 Target Variable Transformation

Since the burnt surface is highly skewed (most fires are small, few are huge), we apply a Logarithmic transformation to stabilize the variance.

```{r}
# Apply Log1p transformation (log(x + 1)) to handle 0 values
wfc_model_final <- wfc_model_final %>%
  mutate(log_surface = log1p(superficie_total_forestal))

# Visualize the effect of the transformation
par(mfrow=c(1,2))
hist(wfc_model_final$superficie_total_forestal, main="Original Surface", col="salmon")
hist(wfc_model_final$log_surface, main="Log-transformed Surface", col="lightblue")
par(mfrow=c(1,1))
```


## 2.4 Data Splitting (Training and Testing)

We split the data into an 80% Training set and a 20% Test set.

```{r}
# Set seed for reproducibility
set.seed(123)

# Create the partition based on the target variable
train_index <- createDataPartition(wfc_model_final$log_surface, p = 0.8, list = FALSE)

# Generate sets
train_set <- wfc_model_final[train_index, ]
test_set  <- wfc_model_final[-train_index, ]

# Print final dimensions
message("Training set rows: ", nrow(train_set))
message("Testing set rows: ", nrow(test_set))
```


# 3. Random Forest Model

```{r}
# Install and load ranger if you haven't
library(ranger)

# Train the Random Forest model
# We predict 'log_surface' using all other columns in 'train_set'
rf_model <- ranger(
  formula         = log_surface ~ ., 
  data            = train_set %>% select(-superficie_total_forestal), # Exclude the original non-log surface
  num.trees       = 500,
  importance      = "permutation", # Important to analyze variable impact later
  seed            = 123
)

# Print model summary
print(rf_model)
```

```{r}
# Get importance
importance_values <- importance(rf_model)
importance_df <- data.frame(
  Variable = names(importance_values),
  Importance = importance_values
) %>% arrange(desc(Importance))

# Plot importance
library(ggplot2)
ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Variable Importance in Wildfire Surface Prediction",
       x = "Predictors", y = "Importance (Permutation)")
```


```{r}
# Predict on test set
predictions <- predict(rf_model, data = test_set)$predictions

# Calculate Performance Metrics (RMSE and R2)
postResample(pred = predictions, obs = test_set$log_surface)
```




# MAP

```{r}
# Load libraries
library(leaflet)
library(dplyr)
library(crosstalk)

# 1. Data Preparation
wfc_map_data <- wfc_final2 %>%
  mutate(
    # Ensure year is a factor for the dropdown selector
    year_factor = as.factor(year),
    # Create the customized popup text in English
    popup_text = paste0(
      "<div style='font-family: Arial; font-size: 12px;'>",
      "<strong>Municipality:</strong> ", municipio, "<br>",
      "<strong>Year:</strong> ", year, "<br>",
      "<strong>Burnt Surface:</strong> ", superficie_total_forestal, " ha<br>",
      "<strong>Land Cover ID:</strong> ", land_cover_id, "<br>",
      "<strong>Cause:</strong> ", causa, "</div>"
    )
  ) %>%
  select(longitude, latitude, year_factor, popup_text, superficie_total_forestal, municipio)

# 2. Create SharedData object
sd <- SharedData$new(wfc_map_data)

# 3. Map layout with single year filter
bscols(
  widths = c(12),
  # Year selector (dropdown)
  filter_select(
    id = "year_selector", 
    label = "Select a specific year to visualize wildfires:", 
    sharedData = sd, 
    group = ~year_factor,
    multiple = FALSE  # Ensures only one year is shown at a time
  ),
  # Leaflet Map with light background
  leaflet(sd, width = "100%", height = 650) %>%
    addTiles() %>% 
    addProviderTiles(providers$CartoDB.Positron) %>% # Clean white/light background
    addCircleMarkers(
      lng = ~longitude, 
      lat = ~latitude,
      radius = ~sqrt(superficie_total_forestal) + 3, 
      color = "#E44D2E",
      stroke = TRUE,
      weight = 1,
      fillOpacity = 0.7,
      popup = ~popup_text,
      label = ~paste(municipio, "-", superficie_total_forestal, "ha") # Hover label
    )
)
```