---
title: "Random Forest"
author: "Gemma Bargalló Solé"
date: "2026-04-28"
html_document:
    code_folding: hide
    toc: true
    toc_float:
      position: left
    toc_depth: 5   
  word_document: default
  pdf_document: default
---

```{r}
library(tidyverse)
library(nanoparquet)
library(here)
library(scales)
library(naniar) 
library(skimr) 
library(patchwork)
library(VIM)
library(gt)
library(purrr)
library(forcats)
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
library(leaflet)
library(dplyr)
library(crosstalk)
library(shiny)
library(shinydashboard)  
library(plotly)
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)
library(ranger)
library(pROC)
```


# 1. Data Import

```{r}
# Final dataset import
wfc_final2 <- read_csv(here("Data", "wfc_final2.csv"))

```
# 2. Quantitative Analysis of Fire Severity: A Random Forest Regression Approach to Burnt Area Prediction.
## 2.1 Pre-processing and Feature Engineering
### 2.1.1 Data Cleaning and Feature Selection
In this step, we filter the columns, handle categorical variables as factors, and prepare the dataset for the model.

```{r}
# Load required libraries
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)

# 1. Feature Selection and Formatting
# FINAL MODEL DATASET SELECTION
wfc_model_final <- wfc_final2 %>%
  select(
    # Target (Change this based on your goal)
    superficie_total_forestal, 
    
    # Location
    altitude_z, latitude, longitude,
    
    # Meteorology
    tx, ppt, thermal_amplitude,
    
    # Time
    month_detected, hour_detected, year,
    
    # Land Cover & Risk
    land_cover_id, afecto_zonas_interfaz_urbano_forestal, 
    afecto_espacio_protegido, afecto_zar,
    
    # Context
    causa
  ) %>%
  # Convert to factors for Random Forest
  mutate(across(where(is.character), as.factor),
         land_cover_id = as.factor(land_cover_id),
         month_detected = as.factor(month_detected),
         # If hour_detected is numeric, we might keep it as is or factorize it
         hour_detected = as.numeric(hour_detected)) %>%
  drop_na()
```


### 2.1.2 Correlation Analysis

We use Spearman correlation because your previous histograms showed that the numerical variables are not normally distributed.

```{r}
# Select only numerical columns for the matrix
numeric_vars <- wfc_model_final %>% select(where(is.numeric))
cor_matrix <- cor(numeric_vars, method = "spearman")

# Plotting the correlation matrix
corrplot(cor_matrix, 
         method = "color", 
         type = "upper", 
         addCoef.col = "black", 
         tl.col = "black", 
         diag = FALSE,
         title = "\nSpearman Correlation Matrix of Predictors",
         mar = c(0,0,2,0))
cor_matrix
```


### 2.1.3 Target Variable Transformation

Since the burnt surface is highly skewed (most fires are small, few are huge), we apply a Logarithmic transformation to stabilize the variance.

```{r}
# Apply Log1p transformation (log(x + 1)) to handle 0 values
wfc_model_final <- wfc_model_final %>%
  mutate(log_surface = log1p(superficie_total_forestal))

# Visualize the effect of the transformation
par(mfrow=c(1,2))
hist(wfc_model_final$superficie_total_forestal, main="Original Surface", col="salmon")
hist(wfc_model_final$log_surface, main="Log-transformed Surface", col="lightblue")
par(mfrow=c(1,1))
```


### 2.1.4 Data Splitting (Training and Testing)

We split the data into an 80% Training set and a 20% Test set.

```{r}
# Define the cutoff point
cutoff <- 2017

# Create the sets chronologically
train_set <- wfc_model_final[wfc_model_final$year <= cutoff, ]
test_set  <- wfc_model_final[wfc_model_final$year > cutoff, ]

# Check dimensions
message("Training set (1998-2017): ", nrow(train_set))
message("Testing set (2018-2022): ", nrow(test_set))

```


## 2.3 Random Forest Model: Regression Approach to Burnt Area Prediction

```{r}
# Load necessary libraries
library(blockCV)
library(sf)
library(ranger)
library(dplyr)

# --- 2. Spatial Cross-Validation Setup ---
# Convert the training set to a spatial object (SF)
# Replace "lon" and "lat" with your actual coordinate column names
train_sf <- st_as_sf(train_set, coords = c("longitude", "latitude"), crs = 4326)

# Create spatial blocks to prevent the model from "memorizing" locations
# This ensures validation happens on areas the model hasn't seen
spatial_folds <- cv_spatial(
  x = train_sf,
  column = "log_surface", 
  k = 10,                  # 10-fold spatial CV
  size = 15000,           # 15km blocks
  selection = "random",
  seed = 123
)

# --- 3. Random Forest Model (Regression) ---
# We use the chronological train_set
# We exclude the original surface and the 'year' column to focus on environmental drivers
rf_model <- ranger(
  formula         = log_surface ~ ., 
  data            = train_set %>% select(-superficie_total_forestal, -year),
  num.trees       = 500,
  importance      = "permutation", 
  seed            = 123
)

print(rf_model)
```


```{r}
# Get importance
importance_values <- importance(rf_model)
importance_df <- data.frame(
  Variable = names(importance_values),
  Importance = importance_values
) %>% arrange(desc(Importance))

# Plot importance
library(ggplot2)
ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Variable Importance in Wildfire Surface Prediction",
       x = "Predictors", y = "Importance (Permutation)")
```



```{r}
# 1. Realitzar prediccions sobre el test_set
predictions <- predict(rf_model, data = test_set %>% select(-superficie_total_forestal, -year))

# 2. Calcular mètriques d'error (RMSE i R-squared real)
actual_values <- test_set$log_surface
predicted_values <- predictions$predictions

# Mètriques
rmse_test <- sqrt(mean((actual_values - predicted_values)^2))
r2_test <- cor(actual_values, predicted_values)^2

message("RMSE on Test Set (2018-2022): ", round(rmse_test, 4))
message("R-squared on Test Set (2018-2022): ", round(r2_test, 4))

# 3. Gràfica de dispersió: Predicció vs Realitat
df_eval <- data.frame(Actual = actual_values, Predicted = predicted_values)

ggplot(df_eval, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.3, color = "darkorange") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Actual vs Predicted Surface (Log Scale)",
       subtitle = "Independent Test Set (2018-2022)",
       x = "Real log(Surface)",
       y = "Predicted log(Surface)")
```

```{r}
# Predict on test set
predictions <- predict(rf_model, data = test_set)$predictions

# Calculate Performance Metrics (RMSE and R2)
postResample(pred = predictions, obs = test_set$log_surface)
```


# 3. Predictive Modeling of Fire Susceptibility: A Binary Classification Framework via Ensemble Learning.
## 3.1 Pre-processing and Feature Engineering
### 3.1.4 Data Splitting (Training and Testing)

- **Threshold definition for severity classification**

Before proceeding with the classification modeling, it is essential to define a threshold that allows us to categorize the fires into two classes: High Severity and Low Severity. Instead of choosing an arbitrary value, we conduct an exploratory analysis of the distribution of the burnt forest area to ensure the statistical robustness of the model.

```{r}
summary(wfc_model_final$superficie_total_forestal)
```
The analysis of the variable total_forest_area reveals a distribution with an extremely high positive skew (right-skewed). While the median is only 0.07 ha, the mean rises to 6.67 ha due to the influence of a few large-scale fires (up to a maximum of 9,301 ha).

This disparity indicates that 75% of the fires (3rd quartile = 0.37 ha) are, in fact, small outbreaks. Therefore, using the arithmetic mean as a cutoff point allows us to identify those events that deviate significantly from the central behavior and represent the real danger for forest management.

```{r}
# Comprovar quants incendis superen la mitjana
table(wfc_final2$superficie_total_forestal > 6.67)
```
The resulting distribution provides a total of 525 cases for the high severity class and 13,872 for the low severity class. This selection allows the model to focus on the 3.6% of the fires with the greatest impact, while also ensuring a sufficiently representative sample of the minority class for the Random Forest algorithm to effectively learn the predictive patterns.


```{r}
# ==========================================
# RANDOM FOREST CLASSIFICATION WITH SPATIAL-TEMPORAL SPLIT
# ==========================================

# 1. Required Libraries
library(caret)
library(blockCV)
library(sf)
library(ranger)
library(pROC)
library(dplyr)

# 2. Data Preparation & Chronological Split
# ------------------------------------------
# Define the binary target variable (Severity)
wfc_model_final$severity <- as.factor(ifelse(wfc_model_final$superficie_total_forestal > 6.67, "High", "Low"))

# Time-based split: 1998-2017 (Train) and 2018-2022 (Test)
cutoff <- 2017
train_set <- wfc_model_final %>% filter(year <= cutoff)
test_set  <- wfc_model_final %>% filter(year > cutoff)

message("Training set (1998-2017): ", nrow(train_set), " rows")
message("Testing set (2018-2022): ", nrow(test_set), " rows")

# 3. Spatial Cross-Validation Setup (blockCV)
# ------------------------------------------
# Convert training set to a spatial object (SF)
# Make sure "longitude" and "latitude" match your column names
train_sf <- st_as_sf(train_set, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

# Generate 20km spatial blocks
spatial_folds <- cv_spatial(
  x = train_sf,
  column = "severity",
  k = 5,
  size = 20000, 
  selection = "random",
  seed = 123
)

# IMPORTANT FIX: Convert blockCV indices to a clean list of integers for caret
# We extract the training indices for each fold
cv_index <- lapply(spatial_folds$folds_list, function(x) as.integer(x$train))

# 4. Model Training Configuration
# ------------------------------------------
fitControl <- trainControl(
  method = "cv",
  index = cv_index,        # Uses our custom spatial blocks
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  sampling = "down"        # Addresses class imbalance
)

# 5. Training the Random Forest (Ranger)
# ------------------------------------------
set.seed(123)
rf_final_model <- train(
  severity ~ tx + ppt + thermal_amplitude + altitude_z + causa + 
             land_cover_id + latitude + longitude + month_detected + 
             hour_detected + afecto_zonas_interfaz_urbano_forestal,
  data = train_set,
  method = "ranger",
  trControl = fitControl,
  metric = "ROC", 
  importance = "permutation"
)

# 6. Evaluation on Independent Test Set (2018-2022)
# ------------------------------------------
# Predictions (Class and Probabilities)
final_preds <- predict(rf_final_model, newdata = test_set)
final_probs <- predict(rf_final_model, newdata = test_set, type = "prob")

# Confusion Matrix for Kappa and Accuracy
print(confusionMatrix(final_preds, test_set$severity))

# AUC-ROC for Discriminative Power
roc_obj <- roc(test_set$severity, final_probs$High)
print(paste("Final Spatial-Temporal AUC:", round(auc(roc_obj), 4)))

# 7. Variable Importance Plot
# ------------------------------------------
plot(varImp(rf_final_model), main = "Variable Importance (Spatial-Temporal RF)")
```



```{r}
# 1. Extraure la importància
importancia_data <- varImp(rf_final_model, scale = FALSE)

# 2. Seleccionar només les, per exemple, 10 més importants
# Això farà que el gràfic sigui molt més net i entenedor
plot(importancia_data, top = 15, main = "Top 10 Drivers of Wildfire Severity", col = "#d95f02")
```

