---
title: "Random Forest"
author: "Gemma Bargalló Solé"
date: "2026-04-28"
html_document:
    code_folding: hide
    toc: true
    toc_float:
      position: left
    toc_depth: 5   
  word_document: default
  pdf_document: default
---

```{r}
library(tidyverse)
library(nanoparquet)
library(here)
library(scales)
library(naniar) 
library(skimr) 
library(patchwork)
library(VIM)
library(gt)
library(purrr)
library(forcats)
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
library(leaflet)
library(crosstalk)
library(shiny)
library(shinydashboard)  
library(plotly)
library(tidyr)
library(corrplot)
library(caret)
library(ranger)
library(pROC)
library(blockCV)
library(sf)

```


# 1. Data Import

```{r}
# Final dataset import
wfc_final2 <- read_csv(here("Data", "wfc_final2.csv"))

```
# 2. Quantitative Analysis of Fire Severity: A Random Forest Regression Approach to Burnt Area Prediction.
## 2.1 Pre-processing and Feature Engineering
### 2.1.1 Data Cleaning and Feature Selection
In this step, we filter the columns, handle categorical variables as factors, and prepare the dataset for the model.

```{r}
# Load required libraries
library(dplyr)
library(tidyr)
library(corrplot)
library(caret)

# 1. Feature Selection and Formatting
# FINAL MODEL DATASET SELECTION
wfc_model_final <- wfc_final2 %>%
  select(
    # Target (Change this based on your goal)
    superficie_total_forestal, 
    
    # Location
    altitude_z, latitude, longitude,
    
    # Meteorology
    tx, ppt, thermal_amplitude,
    
    # Time
    month_detected, hour_detected, year,
    
    # Land Cover & Risk
    land_cover_id, afecto_zonas_interfaz_urbano_forestal, 
    afecto_espacio_protegido, afecto_zar,
    
    # Context
    causa
  ) %>%
  # Convert to factors for Random Forest
  mutate(across(where(is.character), as.factor),
         land_cover_id = as.factor(land_cover_id),
         month_detected = as.factor(month_detected),
         # If hour_detected is numeric, we might keep it as is or factorize it
         hour_detected = as.numeric(hour_detected)) %>%
  drop_na()
```


### 2.1.2 Correlation Analysis

We use Spearman correlation because your previous histograms showed that the numerical variables are not normally distributed.

```{r}
# Select only numerical columns for the matrix
numeric_vars <- wfc_model_final %>% select(where(is.numeric))
cor_matrix <- cor(numeric_vars, method = "spearman")

# Plotting the correlation matrix
corrplot(cor_matrix, 
         method = "color", 
         type = "upper", 
         addCoef.col = "black", 
         tl.col = "black", 
         diag = FALSE,
         title = "\nSpearman Correlation Matrix of Predictors",
         mar = c(0,0,2,0))
cor_matrix
```


### 2.1.3 Target Variable Transformation

Since the burnt surface is highly skewed (most fires are small, few are huge), we apply a Logarithmic transformation to stabilize the variance.

```{r}
# Apply Log1p transformation (log(x + 1)) to handle 0 values
wfc_model_final <- wfc_model_final %>%
  mutate(log_surface = log1p(superficie_total_forestal))

# Visualize the effect of the transformation
par(mfrow=c(1,2))
hist(wfc_model_final$superficie_total_forestal, main="Original Surface", col="salmon")
hist(wfc_model_final$log_surface, main="Log-transformed Surface", col="lightblue")
par(mfrow=c(1,1))
```


### 2.1.4 Data Splitting (Training and Testing)

We split the data into an 80% Training set and a 20% Test set.

```{r}
# Define the cutoff point
cutoff <- 2017

# Create the sets chronologically
train_set <- wfc_model_final[wfc_model_final$year <= cutoff, ]
test_set  <- wfc_model_final[wfc_model_final$year > cutoff, ]

# Check dimensions
message("Training set (1998-2017): ", nrow(train_set))
message("Testing set (2018-2022): ", nrow(test_set))


# --- 2. Creació de Blocs Espacials (Spatial CV) ---
train_sf <- st_as_sf(train_set, coords = c("longitude", "latitude"), crs = 4326)
```


## 2.3 Random Forest Model: Regression Approach to Burnt Area Prediction

```{r}
set.seed(123)
spatial_folds <- cv_spatial(
  x = train_sf,
  column = "log_surface", 
  k = 10,                # 10 folds com deies
  size = 15000,          # Blocs de 15km
  selection = "random"
)

# Extraiem els índexs per a caret
folds_index <- lapply(spatial_folds$folds_list, function(x) as.integer(x[[1]]))
```

## Train model 

```{r}
# --- 3. Configuració de Caret amb índexs espacials ---
fitControl_reg <- trainControl(
  method = "cv",
  index = folds_index,     
  savePredictions = "final"
)

# --- 4. Entrenament amb Ranger ---
set.seed(123)
rf_reg_cv <- train(
  log_surface ~ .,
  data = train_set %>% select(-superficie_total_forestal, -year),
  method = "ranger",
  trControl = fitControl_reg,
  tuneGrid = expand.grid(
    mtry = 3, 
    splitrule = "variance", 
    min.node.size = 5
  ),
  importance = "permutation"
)

print(rf_reg_cv)
```


```{r}
# 1. Extraure la importància i filtrar les 15 millors
importance_df <- data.frame(
  Variable = names(rf_reg_cv$finalModel$variable.importance),
  Importance = rf_reg_cv$finalModel$variable.importance
) %>% 
  arrange(desc(Importance)) %>% 
  slice_head(n = 15)  # Ens quedem només amb el Top 15

# 2. Gràfic amb ggplot2
library(ggplot2)

ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.7) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 15 Variable Importance: Wildfire Surface Prediction",
    subtitle = "Regression Model | 10-fold Spatial Cross-Validation",
    x = NULL, 
    y = "Importance (Permutation)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 10),
    panel.grid.minor = element_blank()
  )
```

## Test Set 2018-2022

```{r}
# --- 5. Predicció sobre el Test Set ---
reg_preds <- predict(rf_reg_cv, newdata = test_set)

# --- 6. Mètriques finals per a la teva taula gt ---
reg_metrics <- postResample(pred = reg_preds, obs = test_set$log_surface)
print(reg_metrics)
```


# 3. Predictive Modeling of Fire Susceptibility: A Binary Classification Framework via Ensemble Learning.
## 3.1 Pre-processing and Feature Engineering
### 3.1.4 Data Splitting (Training and Testing)

- **Threshold definition for severity classification**

Before proceeding with the classification modeling, it is essential to define a threshold that allows us to categorize the fires into two classes: High Severity and Low Severity. Instead of choosing an arbitrary value, we conduct an exploratory analysis of the distribution of the burnt forest area to ensure the statistical robustness of the model.


```{r}
summary(wfc_model_final$superficie_total_forestal)
```
The analysis of the variable total_forest_area reveals a distribution with an extremely high positive skew (right-skewed). While the median is only 0.07 ha, the mean rises to 6.67 ha due to the influence of a few large-scale fires (up to a maximum of 9,301 ha).

This disparity indicates that 75% of the fires (3rd quartile = 0.37 ha) are, in fact, small outbreaks. Therefore, using the arithmetic mean as a cutoff point allows us to identify those events that deviate significantly from the central behavior and represent the real danger for forest management.

```{r}
# Comprovar quants incendis superen la mitjana
table(wfc_final2$superficie_total_forestal > 6.67)
```
The resulting distribution provides a total of 525 cases for the high severity class and 13,872 for the low severity class. This selection allows the model to focus on the 3.6% of the fires with the greatest impact, while also ensuring a sufficiently representative sample of the minority class for the Random Forest algorithm to effectively learn the predictive patterns.

Let's try with another quantile: 

```{r}
llindar_80 <- quantile(wfc_model_final$superficie_total_forestal, 0.80)
message("El percentil 80 correspon a: ", round(llindar_80, 3), " ha")
```
The threshold for 'High Severity' was adjusted from the arithmetic mean to the 80th percentile of the historical distribution. This change is justified by the high skewness of wildfire data and aligns with the INFOCAT plan's criteria, which defines high-risk areas starting at the 0.80 percentile. This approach ensures the model focuses on the top 20% of most impactful events while improving class balance for the Random Forest algorithm."


```{r}
# Comprovar quants incendis superen el threshold
table(wfc_final2$superficie_total_forestal > 0.53)
```
The classification model now focuses on predicting the top 20% of the most severe wildfires (n = 2,878).


```{r}
# RANDOM FOREST CLASSIFICATION: TEMPORAL VALIDATION

# 1. Preparació de les dades i Split Temporal
# ------------------------------------------
wfc_model_final$severity <- as.factor(ifelse(wfc_model_final$superficie_total_forestal > 0.53, "High", "Low"))

train_set <- wfc_model_final %>% filter(year <= 2017)
test_set  <- wfc_model_final %>% filter(year > 2017)

# 2. Downsampling manual del Train Set (Equilibrem 50/50)
# ------------------------------------------------------
set.seed(123)
high_sev_train <- train_set %>% filter(severity == "High")
low_sev_train  <- train_set %>% filter(severity == "Low")

# Igualem el nombre d'incendis petits al de grans
low_sev_balanced <- low_sev_train %>% sample_n(nrow(high_sev_train))
train_balanced <- bind_rows(high_sev_train, low_sev_balanced)

```

```{r}

# 3. Entrenament directe amb Ranger (Més ràpid i sense errors)
# ------------------------------------------------------------
set.seed(123)
rf_final_model <- ranger(
  formula         = severity ~ tx + ppt + thermal_amplitude + altitude_z + causa + 
                    land_cover_id + latitude + longitude + month_detected + 
                    hour_detected + afecto_zonas_interfaz_urbano_forestal,
  data            = train_balanced,
  num.trees       = 500,
  mtry            = 3,            
  importance      = "permutation",
  probability     = TRUE,         
  seed            = 123
)

# 4. Prediccions sobre el Test Set (2018-2022)
# --------------------------------------------
# Obtenim probabilitats i classes
probs_test <- predict(rf_final_model, data = test_set)$predictions
preds_test <- ifelse(probs_test[, "High"] > 0.5, "High", "Low")
preds_test <- factor(preds_test, levels = c("High", "Low"))

# 5. Mètriques de Qualitat 
# -----------------------------------------------
library(caret)
conf_matrix <- confusionMatrix(preds_test, test_set$severity)
print(conf_matrix)

library(pROC)
roc_obj <- roc(test_set$severity, probs_test[, "High"])
message("AUC Final (Temporal Validation 2018-2022): ", round(auc(roc_obj), 4))

```


```{r}
# 1. Extraure la importància i convertir-la en un dataframe net
importancia_df <- data.frame(
  Variable = names(rf_final_model$variable.importance),
  Importance = rf_final_model$variable.importance
) %>% 
  arrange(desc(Importance)) %>% 
  slice_head(n = 15)  # Ens quedem només amb les 15 millors

# 2. Fer el gràfic amb ggplot2 (que queda molt més professional per al TFM)

ggplot(importancia_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "#d95f02") + # El color taronja que t'agrada
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 15 Drivers of Wildfire Severity",
    subtitle = "Classification Model (Threshold: 0.53 ha)",
    x = NULL, 
    y = "Importance (Permutation)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 10)
  )
```

```{r}
library(knitr)
purl("RF.Rmd")
```

